---
editor_options: 
  chunk_output_type: console
---

# Land cover data analysis

In this script, we will analyze land cover changes for multiple land cover classes between 1848 and 2017.  

## Load necessary libraries  
```{r}
library(sf)
library(raster)
library(terra)
library(stars)
library(dplyr)
library(tidyverse)
library(mapview)
library(landscapemetrics)
library(scico)
library(extrafont)
```

## Load the Nilgiris contour

This is the contour to which all other shapefiles will be clipped to and is essentially a 1400 m contour.  

```{r}
# load the 1400m contour
contourNil <- st_read("data/spatial/contour-1400m-nilgiris.shp")
```

## Processing digitized shapefiles from the 1848 map  

We will be loading shapefiles that were digitized by Amrutha Rajan for the 1848 historical map. For an earlier version of the analyses, we had used digitized shapefiles from the year 1870. However, upon further examination, we realized that the the 1870 map was simply a reproduction of the original map from 1848, with missing content.     
```{r}
# list all shapefiles in the directory
nil1848 <- list.files("data/spatial/1848-nilgiris/", full.names = T, recursive = T, pattern=".shp$")

# create vector files
ag1848 <- st_read(nil1848[1]) # type: multipolygon; 6 empty geometries
noData1848 <- st_read(nil1848[2]) # type: multipolygon
plantations1848 <- st_read(nil1848[3]) # type: polygon & multipolygon; 1 empty geometry
roads1848 <- st_read(nil1848[4]) # type: linestring
settlements1848 <- st_read(nil1848[5]) # type: polygon & multipolygon; 3 geometries empty
sholaForest1848 <- st_read(nil1848[6]) # type: polygon & multippolygon; 6 geometries empty  
sholaGrassland1848 <- st_read(nil1848[7]) # type: multipolygon
swamps1848 <- st_read(nil1848[8]) # type: polygon & multipolygon; eight geometries are empty
waterBodies1848 <- st_read(nil1848[9]) # type: multipolygon

# explore and fix any issues with the above vector files
# we need to ensure consistency across files for the sake of merging them into a single geometry collection

# we notice a range of small issues with the shapefiles above
# the geometry type is variable and needs to be consistent
# empty geometries need to be removed
# attribute names need to be consistent across shapefiles

# first, we will remove empty geometries
ag1848 <- ag1848[!st_is_empty(ag1848), ]
noData1848 <- noData1848[!st_is_empty(noData1848), ]
plantations1848 <- plantations1848[!st_is_empty(plantations1848),]
roads1848 <- roads1848[!st_is_empty(roads1848),]
settlements1848 <- settlements1848[!st_is_empty(settlements1848),]
sholaForest1848 <- sholaForest1848[!st_is_empty(sholaForest1848),]
sholaGrassland1848 <- sholaGrassland1848[!st_is_empty(sholaGrassland1848),]
swamps1848 <- swamps1848[!st_is_empty(swamps1848),]
waterBodies1848 <- waterBodies1848[!st_is_empty(waterBodies1848),]

# fixing attribute tables to ensure they are consistent across shapefiles
names(ag1848) <- c("id", "name","geometry")
ag1848$name <- "agriculture"
names(noData1848) <- c("id", "name","geometry")
noData1848$name <- "no_data"
names(plantations1848) <- c("id", "name","geometry")
plantations1848$name <- "plantations"
names(roads1848) <- c("id", "name","geometry")
roads1848$name <- "roads"
names(settlements1848) <- c("id", "name","geometry")
settlements1848$name <- "settlements"
names(sholaForest1848) <- c("id", "name","geometry")
sholaForest1848$name <- "shola_forest"
names(sholaGrassland1848) <- c("id", "name","geometry")
sholaGrassland1848$name <- "shola_grassland"
names(swamps1848) <- c("id", "name","geometry")
swamps1848$name <- "swamps"
names(waterBodies1848) <- c("id", "name","geometry")
waterBodies1848$name <- "water_bodies"

# Note: the roads shapefile is a linestring while the other geometry types are polygon/multipolygon

# transform to UTM 43N
ag1848 <- st_transform(ag1848, 32643)
noData1848 <- st_transform(noData1848, 32643)
plantations1848 <- st_transform(plantations1848, 32643)
roads1848 <- st_transform(roads1848, 32643)
settlements1848 <- st_transform(settlements1848, 32643)
sholaForest1848 <- st_transform(sholaForest1848, 32643)
sholaGrassland1848 <- st_transform(sholaGrassland1848, 32643)
swamps1848 <- st_transform(swamps1848, 32643)
waterBodies1848 <- st_transform(waterBodies1848, 32643)

# creating a single simple feature collection
nil1848 <- rbind(ag1848, plantations1848, settlements1848,
  sholaForest1848, sholaGrassland1848, swamps1848,
  waterBodies1848)

# subsuming swamps under grasslands
nil1848 <- nil1848 %>%
  mutate(name = case_when(
    name == "swamps" ~ "shola_grassland",
    .default = as.character(name)))

# crop the 1848 shapefiles to within the 1400m contour only
nil1848 <- st_buffer(nil1848, dist = 0)
nil1848 <- st_intersection(nil1848, contourNil)
nil1848 <- nil1848[,-c(3:5)]

# Create a common boundary for clipping files from other time periods (see below)
all1848 <- st_buffer(st_union(nil1848), dist = 0)
```


## Load the shapefiles from 1973, 1995 and 2017 for the Nilgiris  
```{r}
# load the 1973, 1995 and 2017 shapefiles
nil1973 <- st_read("data/spatial/1973-nilgiris/1973.shp")
nil1995 <- st_read("data/spatial/1995-nilgiris/1995.shp")
nil2017 <- st_read("data/spatial/2017-nilgiris/2017.shp")

# Each of the above shapefiles have slightly different characteristics
# class name is present only in nil2017 and not in the others
# nil 1973 and nil 1995 are of geometry type MULITPOLYGON while nil 2017 is of geometry type POLYGON

# Let's first add class name to nil1973 shapefile
nil1973 <- nil1973 %>%
  mutate(class = case_when(
    gridcode == 1 ~ "Shola Grassland",
    gridcode == 2 ~ "Shola Forest",
    gridcode == 3 ~ "Timber Plantations",
    gridcode == 4 ~ "Tea Plantations",
    gridcode == 5 ~ "Ochlandra",
    gridcode == 6 ~ "Settlements",
    gridcode == 7 ~ "Agricultural Land",
    gridcode == 8 ~ "Water bodies"
  ))

# Let's add class name to nil1995 shapefile
nil1995 <- nil1995 %>%
  mutate(class = case_when(
    gridcode == 1 ~ "Shola Grassland",
    gridcode == 2 ~ "Shola Forest",
    gridcode == 3 ~ "Timber Plantations",
    gridcode == 4 ~ "Tea Plantations",
    gridcode == 5 ~ "Ochlandra",
    gridcode == 6 ~ "Settlements",
    gridcode == 7 ~ "Agricultural Land",
    gridcode == 8 ~ "Water bodies"
  ))

# crop the 1973, 1995 and 2017 shapefiles to within the 1400m contour only
nil1973 <- st_intersection(nil1973, all1848)
nil1995 <- st_intersection(nil1995, all1848)
nil2017 <- st_intersection(nil2017, all1848)
```

## Rasterization of vector files

The decision to use a vector or raster for further analysis depends on the question being asked. Please refer to script in code/01_vectorVsRaster.R
```{r}
# nil11848 raster
# scale: 2000ft to 1 inch (1:24000; 12 metres resolution)
vectnil1848 <- terra::vect(nil1848)
emptyRast <- terra::rast(res = 6, xmin = 660771, xmax = 718316.3, ymin = 1240554, ymax = 1275492, crs = "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs")
rastnil1848 <- terra::rasterize(vectnil1848, emptyRast, "name")

# Let's load the 1973, 1995 and 2017 series and rasterize them
vectnil1973 <- terra::vect(nil1973)
emptyRast1973 <- terra::rast(res = 30, xmin = 660771, xmax = 718316.3, ymin = 1240554, ymax = 1275492, crs = "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs")
rastnil1973 <- terra::rasterize(vectnil1973, emptyRast1973, "class")
sz1973 <- cellSize(rastnil1973, unit = "m")
areaRast1973 <- zonal(sz1973, rastnil1973, sum)
areaRast1973SqKm <- (areaRast1973$area / 1000000)
areaRast1973 <- cbind(areaRast1973, areaRast1973SqKm)
names(areaRast1973) <- c("class", "areaInMeters", "sumArea")

# 1995
vectnil1995 <- vect(nil1995)
emptyRast1995 <- terra::rast(res = 30, xmin = 660771, xmax = 718316.3, ymin = 1240554, ymax = 1275492, crs = "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs")
rastnil1995 <- terra::rasterize(vectnil1995, emptyRast1995, "class")
sz1995 <- cellSize(rastnil1995, unit = "m")
areaRast1995 <- zonal(sz1995, rastnil1995, sum)
areaRast1995SqKm <- (areaRast1995$area / 1000000)
areaRast1995 <- cbind(areaRast1995, areaRast1995SqKm)
names(areaRast1995) <- c("class", "areaInMeters", "sumArea")

# 2017
vectnil2017 <- vect(nil2017)
emptyRast2017 <- terra::rast(res = 30, xmin = 660771, xmax = 718316.3, ymin = 1240554, ymax = 1275492,  crs = "+proj=utm +zone=43 +datum=WGS84 +units=m +no_defs")
rastnil2017 <- terra::rasterize(vectnil2017, emptyRast2017, "class")
sz2017 <- cellSize(rastnil2017, unit = "m")
areaRast2017 <- zonal(sz2017, rastnil2017, sum)
areaRast2017SqKm <- (areaRast2017$area / 1000000)
areaRast2017 <- cbind(areaRast2017, areaRast2017SqKm)
names(areaRast2017) <- c("class", "areaInMeters", "sumArea")

# resampling rasters to ensure they are all of the same spatial resolution and extent
# We have to do this for the nil1848 raster

resam1848 <- terra::resample(rastnil1848, rastnil2017, method = "near")
sz1848 <- cellSize(resam1848, unit = "m")
areaNil1848 <- zonal(sz1848, resam1848, sum)
areaNil1848SqKm <- (areaNil1848$area / 1000000)
areaNil1848 <- cbind(areaNil1848, areaNil1848SqKm)
```

## Write the shapefiles and rasters to file
```{r}
# vectors
st_write(nil1848, "results/cropped-shapefiles/1848.shp", driver = "ESRI Shapefile")
st_write(nil1973, "results/cropped-shapefiles/1973.shp", driver = "ESRI Shapefile")
st_write(nil1995, "results/cropped-shapefiles/1995.shp", driver = "ESRI Shapefile")
st_write(nil2017, "results/cropped-shapefiles/2017.shp", driver = "ESRI Shapefile")

# rasters
terra::writeRaster(resam1848, "results/cropped-rasters/1848.tif")
terra::writeRaster(rastnil1973, "results/cropped-rasters/1973.tif")
terra::writeRaster(rastnil1995, "results/cropped-rasters/1995.tif")
terra::writeRaster(rastnil2017, "results/cropped-rasters/2017.tif")
```

## Cleaning up the attributes for visualization
```{r}
## Note, since there was no distinction made between tea and timber plantations for the 1848 map, we will subsume these two classes under a single class - plantations (for the sake of calculating areas)

areaRast1973 <- areaRast1973 %>%
  mutate(class = case_when(
    class == "Agricultural Land" ~ "agriculture",
    class == "Settlements" ~ "settlements",
    class == "Shola Forest" ~ "shola_forest",
    class == "Shola Grassland" ~ "shola_grassland",
    class == "Tea Plantations" ~ "plantations",
    class == "Timber Plantations" ~ "plantations",
    class == "Water bodies" ~ "water_bodies"))

areaNil1973 <- areaRast1973 %>% 
  group_by(class) %>% 
  summarise(areaInMeters = sum(areaInMeters),
         sumArea = sum(sumArea)) %>% ungroup()

areaRast1995 <- areaRast1995 %>%
  mutate(class = case_when(
    class == "Agricultural Land" ~ "agriculture",
    class == "Settlements" ~ "settlements",
    class == "Shola Forest" ~ "shola_forest",
    class == "Shola Grassland" ~ "shola_grassland",
    class == "Tea Plantations" ~ "plantations",
    class == "Timber Plantations" ~ "plantations",
    class == "Water bodies" ~ "water_bodies"))

areaNil1995 <- areaRast1995 %>% 
  group_by(class) %>% 
  summarise(areaInMeters = sum(areaInMeters),
         sumArea = sum(sumArea)) %>% ungroup()

areaRast2017 <- areaRast2017 %>%
  mutate(class = case_when(
    class == "Agricultural Land" ~ "agriculture",
    class == "Settlements" ~ "settlements",
    class == "Shola Forest" ~ "shola_forest",
    class == "Shola Grassland" ~ "shola_grassland",
    class == "Tea Plantations" ~ "plantations",
    class == "Timber Plantations" ~ "plantations",
    class == "Water bodies" ~ "water_bodies"))

areaNil2017 <- areaRast2017 %>% 
  group_by(class) %>% 
  summarise(areaInMeters = sum(areaInMeters),
         sumArea = sum(sumArea)) %>% ungroup()
```

### Calculating areas of different land cover classes across time periods using rasters

Making plots of the overall area of land cover types
```{r}
# Joining all dataframes to create a single one for plotting and visualization
names(areaNil1848) <- c("class", "areaInMeters","sumArea")

areaCalc <- purrr::reduce(list(
  areaNil1848, 
  areaNil1973, 
  areaNil1995, 
  areaNil2017
), dplyr::full_join, by = "class")

names(areaCalc) <- c("class", "areaMt1848", "1848", "areaMt1973", "1973", "areaMt1995", "1995", "areaMt2017", "2017")

areaCalc <- areaCalc %>%
  select("class", "1848", "1973", "1995", "2017") %>%
  pivot_longer(!class, names_to = "Year", values_to = "Area")

write.csv(areaCalc, "results/totalArea-by-landCover-timePeriod.csv", row.names = F)

# make plot
fig_area <- ggplot(areaCalc, aes(x = class, 
                                 y = Area, fill = class)) + 
  geom_bar(stat = "identity", 
           position = position_dodge()) + 
  scale_fill_manual(values = c('#b35806', '#d73027','#762a83',
                   '#1b7837', '#7fbf7b','#4575b4'))+
    geom_text(aes(label = round(Area), hjust = "middle", 
                vjust = -0.5), family = "Century Gothic",
    position = position_dodge(), angle = 0, 
    size = 5) +
  facet_wrap(~Year) +
  theme_bw() +
  labs(
    x = "\nLand cover type",
    y = "Area in sq.km. \n"
  ) +
  theme(text = element_text(size=14,  family="Century Gothic"),
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"),
    axis.text = element_text(family = "Century Gothic", 
                             size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, 
                               hjust = 1),
    legend.position = "none")

ggsave(fig_area,
  filename = "figs/fig_totalArea_landCover.png", width = 15, height = 13, device = png(), units = "in", dpi = 600)
dev.off()
```

### Create a figure of proportion of landcover across areas over time 
```{r}
# get percentArea occupied by each land cover class
percentArea <- areaCalc %>%
  group_by(Year) %>%
  mutate(
    totalArea = sum(Area, na.rm = T),
    percentArea = (Area / totalArea) * 100
  )

# plot figure
fig_percent_area <- ggplot(percentArea, aes(x = class, y = percentArea, fill = class)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = c('#b35806', '#d73027','#762a83',
                   '#1b7837', '#7fbf7b','#4575b4')) +
  geom_text(aes(label = round(percentArea, digits = 1), hjust = "middle", vjust = -0.5), position = position_dodge(), family = "Century Gothic", angle = 0, size = 5) +
  facet_wrap(~Year) +
  theme_bw() +
  labs(
    x = "\nLand cover type",
    y = "Percent Area \n"
  ) +
  theme(text = element_text(size=14,  family="Century Gothic"),
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),axis.text = element_text(family = "Century Gothic", 
                               size = 14),
    axis.text.x = element_text(angle = 90, 
                               vjust = 0.5, hjust = 1),
    legend.position = "none")

ggsave(fig_percent_area, filename = "figs/fig_percentArea_landCover.png", width = 14, height = 13, device = png(), units = "in", dpi = 500)
dev.off()
```

