---
editor_options: 
  chunk_output_type: console
---

# Exploratory analysis of occurrence data

This script carries out exploratory analysis of historical and modern occurrence data from the Nilgiri hills. In this script, we plot the total count of specimens by time period and the number of specimens per species by time period.   

### Load necessary libraries
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
```

### Load the historical occurrence data and explore patterns
```{r}
hist_occ <- read.csv("data/historical-occurrence-data.csv")

# count of data by year
occ_year <- hist_occ %>%
  group_by(year) %>%
  count()

# The above count suggests that majority of the bird specimens were recorded in 1881, followed by some data from 1876.

# Let us create a new column that bins year by time period. For the sake of exploratory analysis, we bin data into 50 year time periods (1850-1900, 1900-1950, 1950-2000, post2000)

hist_occ <- hist_occ %>%
  filter(year <= 1900)
```

### Examining count of specimens by locality
```{r}
occ_locality <- hist_occ %>%
  group_by(historical_site_code) %>%
  count()

fig_loc_timePeriod <- ggplot(occ_locality, aes(x = historical_site_code, y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nLocality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_loc_timePeriod, filename = "figs/fig_histCount_locality.png", width = 12, height = 7, device = png(), units = "in", dpi = 300)
dev.off()
```

### Examining species specific counts by locality
```{r}
## species count
spp_count <- hist_occ %>%
  group_by(scientific_name) %>%
  count() 

unique(spp_count$scientific_name) # a total of 147 species recorded between 1850 to 1900

# filter observations (identify how many species were not singletons, a min of 4)
spp_minFourObs <- spp_count %>%
  filter(n >= 4) # only 68 species had more than or equal to 4 observations in the historical dataset

fig_spp <- ggplot(spp_count, aes(x = scientific_name, y = n, fill = scientific_name)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nSpecies",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_spp, filename = "figs/fig_histSpp.png", width = 50, height = 20, device = png(), units = "in", dpi = 300, limitsize = F)
dev.off()

# Examining species by locality
# Plotting locations that had a min of 4 species observations 
spp_loc <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  filter(n >= 4)

fig_spp_loc <- ggplot(spp_loc, aes(x = scientific_name, y = n, fill = scientific_name)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  facet_wrap(~historical_site_code) +
  theme_bw() +
  labs(
    x = "\nSpecies & Locality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_spp_loc, filename = "figs/fig_histSpp_locality_min4.png", width = 20, height = 12, device = png(), units = "in", dpi = 300)
dev.off()
```

### Load modern occurrence data and explore patterns
```{r}
mod_occ <- read.csv("data/modern-occurrence-data.csv") %>%
  filter(!is.na(historical_site_code)) %>%
  filter(!is.na(species_code))

# total observations of species in the modern surveys
modOcc_spp <- mod_occ %>%
  group_by(species_code) %>%
  count() # 2301 observations of 102 bird species

## only include species that are present in the historical dataset
mod_occ <- mod_occ%>%
  filter(scientific_name %in% hist_occ$scientific_name)

# plot figure of species by locality
mod_occ_loc <- mod_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

# figure of species by locality
fig_modSpp_loc <- ggplot(mod_occ_loc, aes(x = historical_site_code, y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  facet_wrap(~scientific_name) +
  theme_bw() +
  labs(
    x = "\nSpecies and Locality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_modSpp_loc, filename = "figs/fig_modSpp_loc.png", width = 40, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## Some interesting patterns of certain species being more abundant/detected in certain sites over others.
```

