---
editor_options: 
  chunk_output_type: console
---

# Exploratory analysis of occurrence data

This script carries out exploratory analysis of historical and modern occurrence data from the Nilgiri hills. In this script, we plot the total count of specimens by time period and the number of specimens per species by time period.   

### Load necessary libraries
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
```

### Load the historical occurrence data and explore patterns
```{r}
hist_occ <- read.csv("data/historical-occurrence-data.csv")

# count of data by year
occ_year <- hist_occ %>%
  group_by(year) %>%
  count()

# The above count suggests that majority of the bird specimens were recorded in 1881, followed by some data from 1876.

# Let us create a new column that bins year by time period. For the sake of exploratory analysis, we bin data into 50 year time periods (1850-1900, 1900-1950, 1950-2000, post2000)

hist_occ <- hist_occ %>%
  filter(year <= 1900)
```

### Examining count of specimens by locality
```{r}
# ten unique historical site localities across which species were recorded
occ_locality <- hist_occ %>%
  group_by(historical_site_code) %>%
  count()

# Examining species by locality
spp_loc <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

# how many unique sites
prop_count <- spp_loc %>%
  group_by(scientific_name) %>%
  summarise(uniqueSites = length(unique(historical_site_code)))

# moving forward, we only choose species that occur across
# a minimum of 30% of all sites
prop_count <- prop_count %>%
  filter(uniqueSites >=3)

## The above filter leaves with a total of 49 species.

## we further subset the above data to include only species that had a minimum species abundance data of over five

## choosing only species that occurred across 30% of sites
hist_occ <- hist_occ %>%
  filter(scientific_name %in% prop_count$scientific_name)

## further filtering species to only include species that had a minimum number of species records of five
spp_count <- hist_occ %>%
  group_by(scientific_name) %>%
  count() %>%
  filter(n >=5)

unique(spp_count$scientific_name) # a total of 45 species recorded between 1850 to 1900 that occurred in atleast

# adding an additional filter
# we now have species that occurred in atleast 30% of sites and
# had a minimum historical specimen abundance of five

hist_occ <- hist_occ %>%
  filter(scientific_name %in% spp_count$scientific_name)

# testing
unique(hist_occ$scientific_name)
```

## Visualizations of historical data
```{r}
fig_loc_timePeriod <- ggplot(occ_locality, aes(x = historical_site_code, y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nLocality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_loc_timePeriod, filename = "figs/fig_histCount_locality.png", width = 12, height = 7, device = png(), units = "in", dpi = 300)
dev.off()
```

### Examining species specific counts by locality
```{r}
fig_spp <- ggplot(spp_count, aes(x = reorder(scientific_name, -n), y = n, fill = scientific_name)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nSpecies",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_spp, filename = "figs/fig_histSpp.png", width = 50, height = 20, device = png(), units = "in", dpi = 300, limitsize = F)
dev.off()

## Re-examining species by locality after filtering 
spp_loc <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

fig_spp_loc <- ggplot(spp_loc, aes(x = reorder(scientific_name,-n), y = n, fill = scientific_name)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  facet_wrap(~historical_site_code) +
  theme_bw() +
  labs(
    x = "\nSpecies & Locality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_spp_loc, filename = "figs/fig_histSpp_locality.png", width = 20, height = 12, device = png(), units = "in", dpi = 300)
dev.off()
```

### Load modern occurrence data and explore patterns
```{r}
mod_occ <- read.csv("data/modern-occurrence-data.csv") %>%
  filter(!is.na(historical_site_code)) %>%
  filter(!is.na(species_code))

# total observations of species in the modern surveys
modOcc_spp <- mod_occ %>%
  group_by(species_code) %>%
  count() # 2301 observations of 102 bird species

## only include species that are present in the historical dataset
mod_occ <- mod_occ%>%
  filter(scientific_name %in% hist_occ$scientific_name)

# plot figure of species by locality
mod_occ_loc <- mod_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

# figure of species by locality
fig_modSpp_loc <- ggplot(mod_occ_loc, aes(x = reorder(historical_site_code,-n), y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  facet_wrap(~scientific_name) +
  theme_bw() +
  labs(
    x = "\nSpecies and Locality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_modSpp_loc, filename = "figs/fig_modSpp_loc.png", width = 40, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## Some interesting patterns of certain species being more abundant/detected in certain sites over others.
```

## write the subset data to file
```{r}
write.csv(hist_occ, "results/histSubset.csv", row.names = F)
write.csv(mod_occ, "results/modSubset.csv", row.names = F)
```

