---
editor_options: 
  chunk_output_type: console
---

# Exploratory analysis of occurrence data

This script carries out exploratory analysis of historical and modern occurrence data from the Nilgiri hills.   

### Load necessary libraries
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
```

### Load the historical occurrence data and explore patterns

The historical occurrence data is a combination of data from museum specimens and those observations made in published literature/journals (titled journal data). 
```{r}
hist_occ <- read.csv("data/historical-occurrence-data.csv")

# after exploring the data, we will apply the following filters on the dataset

# a: include only data until 1950
filt_hist <- hist_occ %>%
  filter(year <= 1950)

# b: remove localities with limited amount of data
# here we exclude localities that had less than 10 cumulative observations/records between 1850-1950
# localities to remove include: 
# BGOT, EMVL, KNSL, KUKL, LVDL, RNMD, TMTE
filt_hist <- filt_hist %>%
  filter(!historical_site_code %in% c("BGOT","EMVL",
                                      "KNSL",
                                      "KUKL","LVDL",
                                   "RNMD","TMTE"))

# c: remove specific species


# count of data by year
occ_year <- filt_hist %>%
  group_by(year, historical_site_code) %>%
  count()

spp_count <- trial %>%
  group_by(scientific_name) %>%
  count()

collector <- hist_occ %>%
  group_by(collectedBy, year) %>%
  count()

# The above count suggests that majority of the bird specimens were recorded in 1881, followed by some data from 1876.

# For further downstream analysis, we ignore data from 1901 onwards
hist_occ <- hist_occ %>%
  filter(year <= 1901)
```

### Examining count of specimens by locality
```{r}
# ten unique historical site localities across which species were recorded
occ_locality <- hist_occ %>%
  group_by(historical_site_code) %>%
  count()

# Examining species by locality
spp_loc <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

# how many unique sites
prop_count <- spp_loc %>%
  group_by(scientific_name) %>%
  summarise(uniqueSites = length(unique(historical_site_code)))

## count unique species list
spp_count <- hist_occ %>%
  group_by(scientific_name) %>%
  count()

unique(spp_count$scientific_name) # a total of 148 species were collected/recorded between 1850 and 1901 across 12 unique historical localities identified
```

## Visualizations of historical data
```{r}
fig_loc_timePeriod <- ggplot(occ_locality, aes(x = reorder(historical_site_code,-n), y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nLocality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_loc_timePeriod, filename = "figs/fig_histCount_locality.png", width = 12, height = 7, device = png(), units = "in", dpi = 300)
dev.off()
```


## Examining species specific counts by locality
```{r}
fig_spp <- ggplot(spp_count, aes(x = reorder(scientific_name, -n), y = n, fill = scientific_name)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  theme_bw() +
  labs(
    x = "\nSpecies",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 14),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_spp, filename = "figs/fig_histSpp.png", width = 50, height = 20, device = png(), units = "in", dpi = 300, limitsize = F)
dev.off()
```

## Load modern occurrence data and explore patterns
```{r}
mod_occ <- read.csv("data/modern-occurrence-data.csv") %>%
  filter(!is.na(historical_site_code)) %>%
  filter(!is.na(species_code)) %>%
  filter(scientific_name %in% spp_count$scientific_name)

# total observations of species in the modern surveys
modOcc_spp <- mod_occ %>%
  group_by(species_code) %>%
  count() # 2301 observations of 102 bird species

# plot figure of species by locality
mod_occ_loc <- mod_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count()

# figure of species by locality
fig_modSpp_loc <- ggplot(mod_occ_loc, aes(x = reorder(historical_site_code,-n), y = n, fill = historical_site_code)) +
  geom_bar(stat = "identity", position = position_dodge(), fill = "#883107", alpha = 0.9) +
  geom_text(aes(label = n, hjust = "middle", vjust = -0.5),
    position = position_dodge(), angle = 0, size = 5
  ) +
  facet_wrap(~scientific_name) +
  theme_bw() +
  labs(
    x = "\nSpecies and Locality",
    y = "Count\n"
  ) +
  theme(
    axis.title = element_text(
      family = "Century Gothic",
      size = 14, face = "bold"
    ),
    axis.text = element_text(family = "Century Gothic", size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

ggsave(fig_modSpp_loc, filename = "figs/fig_modSpp_loc.png", width = 40, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## Some interesting patterns of certain species being more abundant/detected in certain sites over others.
```

