---
editor_options: 
  chunk_output_type: console
---

# Species relative abundances over time

In this script, we calculate species relative abundances over time by replicating a novel approach proposed by Gotelli et al. 2021 (who provides a methodological approach to calculate relative species abundances using historical museum records).

The assumption is that the level of the analysis is at the 'historic-site' and not the 'modern-site' (which essentially corresponds to multiple unique modern survey sites for every 'historic-site'). 

## Load necessary libraries
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(lattice)
library(data.table)
library(ggrepel)
library(report)
library(lme4)
library(glmmTMB)
library(multcomp)
library(ggstatsplot)
library(paletteer)
library(ggpubr)
library(goeveg)
library(sjPlot)
library(patchwork)

# source custom functions
source("code/02_relative-species-abundance-functions.R")
```

## Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.  
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster::raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

## For the historical occurrence data, we will choose data from a single time period: 1850-1900
```{r}
## read in historical occurrence data
hist_occ <- read.csv("data/historical-occurrence-data.csv")

## filter data
hist_occ <- hist_occ %>%
  filter(year <= 1901)

# calculate total count of historical data at each site
hist_occ <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  summarise("1850-1900" = sum(n)) 

# left join the sites dataframe
hist_occ <- left_join(hist_occ, sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) 

## read in modern occurrence data
mod_occ <- read.csv("data/modern-occurrence-data.csv")

## group by scientific name and modern site code
mod_occ <- mod_occ %>%
  filter(!is.na(historical_site_code)) %>%
  filter(!is.na(scientific_name)) %>%
  group_by(scientific_name, modern_site_code) %>%
  summarise("2021" = sum(number))

## left join the sites dataframe
mod_occ <-  left_join(mod_occ, sites, by=c("modern_site_code"="modern_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)
```

## Calculate dirichlet distributions of relative abundance

Here, we will functions from Gotelli et al. 2021 to calculate relative species abundances between historic and modern periods.  
```{r}
# note: the level of the analysis is at the 'historic-site' and not the 'modern-site' (which essentially corresponds to two or three unique sites for every 'historic-site')
# to ensure that the analysis is consistent, we sum abundances across the modern sites corresponding to each historic site
# However, our analysis calculates relative species abundances for the give time period rather than by site + time-period

## calculating relative species abundances for the time period 1850-1900

# we remove modern site_code and get only distinct rows
unique_hist <- hist_occ[,c(1,2,3)] %>%
  group_by(scientific_name, historical_site_code) %>%
  distinct() %>%
  group_by(scientific_name) %>%
  summarise("relAbund1850" = sum(`1850-1900`)) 

# we remove modern site_code and get only distinct rows
unique_mod <- mod_occ[,c(1,3,5)] %>%
  group_by(scientific_name, historical_site_code) %>%
  distinct() %>%
  group_by(scientific_name) %>%
  summarise("relAbund2021" = sum(`2021`)) 

hist_mod <- full_join(unique_hist, unique_mod) %>%
  replace_na(list(relAbund1850 = 0, relAbund2021 = 0)) 

## above dataframe suggests that there are 69 species that had historical records/observations, but were not detected in the modern resurvey
## similarly, there are 23 species unique to the modern resurvey that were not recorded/collected in the historical time period

# apply dirichlet distribution to the historical data
dat <- tibble::deframe(hist_mod[,1:2]) ## choosing 1850-1900 data

relAbun1850 <- dirch_stats(dat) %>%
  dplyr::select(species, bayes_mean) %>%
  rename(., bayes_mean_1850 = bayes_mean)

# apply dirichlet distribution to the modern survey
dat <- tibble::deframe(hist_mod[,c(1,3)]) ## choosing modern data

relAbun2021 <- dirch_stats(dat) %>%
  dplyr::select(species, bayes_mean) %>%
  rename(., bayes_mean_2021 = bayes_mean)

### join the relative abundance datasets together
relAbun <- full_join(relAbun1850, relAbun2021, by = c("species")) %>%
  #replace(is.na(.), 0) %>%
  arrange(species) %>%
  rename(., `1850-1900` = "bayes_mean_1850") %>%
  rename(., `2021` = "bayes_mean_2021") %>%
  rename(., scientific_name = species)
```

## Visualizing changes in relative abundance over time

```{r}
fig_1850v2021_rel <- ggplot(data=relAbun,aes(x=`2021`,y=`1850-1900`)) +
   scale_y_log10() +
  geom_abline(slope=1, intercept=0,linetype="dashed") +
  scale_x_log10() +
  labs(y = "1850-1900 Relative abundance", 
       x = "2021 Relative abundance") +
   geom_point(shape = 21, colour = "black", fill = "white", size = 2, stroke = 1)+
   geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95,linetype="solid")+
  stat_regline_equation(aes(label = ..rr.label..)) +
  theme_bw() +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25))

ggsave(fig_1850v2021_rel, filename = "figs/fig_1850-1900vs2021_relAbun.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Boxplot visualization of overall relative abundance between timeperiods
```{r}
data_for_boxplot <- relAbun %>%
  dplyr::select(-scientific_name) %>%
  pivot_longer(everything())

fig_relAbun_over_time <- ggbetweenstats(
  data = data_for_boxplot,
  x = name,
  y = value,
  xlab = "Time Period", 
  ylab = "Relative Abundance",
  title = "Relative abundance by time period",
  plot.type = "box",
  pairwise.comparisons = T) +
  scale_y_log10() +
  scale_color_manual(values = c("#9EB0FFFF", "#122C39FF","#D5857DFF")) +   theme(plot.title = element_text(family = "Century Gothic",
      size = 18, face = "bold"),
    axis.title = element_text(family = "Century Gothic",
      size = 16, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      plot.subtitle = element_text(
      family = "Century Gothic", 
      size = 14, 
      face = "bold",
      color="#1b2838"
    ))

ggsave(fig_relAbun_over_time, filename = "figs/fig_relAbun_landscapeLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Species relative abundances at the site-level  

We now recalculate species relative abundances at the site-level, which will be used for downstream comparisons with change in land cover and climate at a given site. 
```{r}
# historical data
## we remove modern site_code and get only distinct rows
hist_site_level <- hist_occ[,c(1:3)] %>%
  distinct(.)


    count(scientific_name, as.factor(historical_site_code),
        .drop = FALSE)
  
##  modern resurvey
## analysis is being done at the level of the historical site code
mod_site_level <- mod_occ[,c(1,3,5)] %>%
  group_by(scientific_name, historical_site_code) %>%
  mutate("2021" = sum(`2021`)) %>%
  distinct(.)

## join data frames together
site_level_relAbun <- full_join(hist_site_level,
                                mod_site_level) %>%
  replace_na(list(`1850-1900` = 0, `2021` = 0)) %>% 
  ungroup() %>%
  complete(scientific_name, historical_site_code, fill = list(`1850-1900` = 0, `2021` = 0))

## site-level historical relative abundance
relSpAbund_hist <- c()

for(i in 1:length(unique(site_level_relAbun$historical_site_code))) {
  
  # for every unique historical site
  a <- site_level_relAbun %>%
  filter(historical_site_code == unique(site_level_relAbun$historical_site_code)[i])
  
  # subset historical data
  b <- tibble::deframe(a[,c(1,3)]) ## choosing 1850-1900 data
  
  loc <- unique(site_level_relAbun$historical_site_code)[i] # what's the site_code?
  
  ## historical relative abundance
  relSpAbund_hist[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","1850", sep = ""))

  relSpAbund_hist[[i]] <- relSpAbund_hist[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_hist)[i] <- loc
}

# save object but full_join across lists
relSpAbund_hist <- relSpAbund_hist %>%
    reduce(.f=full_join)

# modern resurvey relative abundance - site level
relSpAbund_2021 <- c()

for(i in 1:length(unique(site_level_relAbun$historical_site_code))) {
  
  a <- site_level_relAbun %>%
  filter(historical_site_code == unique(site_level_relAbun$historical_site_code)[i])

  b <- tibble::deframe(a[,c(1,4)]) ## choosing modern data
  
  loc <- unique(site_level_relAbun$historical_site_code)[i] # what's the site_code?
  
  relSpAbund_2021[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","2021", sep = ""))

  relSpAbund_2021[[i]] <- relSpAbund_2021[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_2021)[i] <- loc 
}

# save object but full_join across lists
relSpAbund_2021 <- relSpAbund_2021 %>%
    reduce(.f=full_join)

## join all the datasets together
relSpAbund_by_site <- full_join(relSpAbund_hist, relSpAbund_2021, by = "scientific_name") %>%
  arrange(scientific_name) 
```

## Visualizing differences in relative abundance over time at the site level
```{r}
## preparing dataframe for visualization
relSpAbund <- relSpAbund_by_site %>%
  pivot_longer(!scientific_name, names_to = "site_year", values_to = "relAbundance") %>%
  separate(site_year, into = c("site", "year"), sep = "_") 

## Please note: for the sake of downstream analysis, we have renamed 1850-1900 as 1850. In other words, each year - except for 2021 represents a 50 year time period. 

fig_relAbun_siteLevel <- relSpAbund %>%
  dplyr::select(-scientific_name) %>%
  ggplot(aes(x=year, y = relAbundance,
             fill = year)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~site)+
  xlab("Time period") +
  ylab("Relative abundance") +
  scale_fill_scico_d(palette = "roma") +
  theme_bw() +
  theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14)) 

ggsave(fig_relAbun_siteLevel, filename = "figs/fig_relAbun_siteLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Writing results to file
```{r}
write.csv(relAbun, "results/species-relative-abundance.csv", row.names = F)
write.csv(relSpAbund, "results/species-relative-abundance-siteLevel.csv", row.names = F)
```

