---
editor_options: 
  chunk_output_type: console
---

### Species relative abundances over time

We replicate analyses following Gotelli et al. 2021 (who provides a methodological approach to calculate relative species abundances using historical museum records) and Gómez et al. 2021 who assumed two scenarios while calculating extinctions and colonizations - "The first is based on the assumption that all surveys accurately describe the avifauna present and that any records of species not detected in prior surveys of San Antonio represent real colonizations. The second scenario is based on the assumption that all the novel species (i.e., those detected for the first time after the 1910s) were false absences."

The second assumption is that the level of the analysis is at the 'historic-site' and not the 'modern-site' (which essentially corresponds to multiple unique modern survey sites for every 'historic-site'). While calculating relative abundances and for estimation of turnover and to ensure that the analysis is consistent, we sum abundances across the modern sites corresponding to each historic site.  

### Load necessary libraries
```{r}
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)
library(codyn)

# source custom functions
source("code/02_relative-species-abundance-functions.R")
```

### Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.  
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- extract(alt, sites)
sites <- cbind(sites, elev)
```


For the historical occurrence data, we will choose data from three time periods: 1850-1900, 1900-1950 and 1950-2000. 
```{r}
## read in historical occurrence data
hist_occ <- read.csv("data/historical-occurrence-data.csv")

### pre-1900
hist_occ_pre1900 <- hist_occ %>%
  filter(year <= 1900)

### fix the count column
hist_occ_pre1900 <-  hist_occ_pre1900 %>%
  mutate(histCount = case_when(individualCount == "X" ~ 1,
                               individualCount == "" ~ 1,
                               TRUE ~ as.numeric(individualCount)))

# calculate total abundance of historical data at each site
hist_occ_count_pre1900 <- hist_occ_pre1900 %>%
  group_by(scientific_name, historical_site_code) %>%
  summarise("1850-1900" = sum(histCount)) 

# left join the sites dataframe
hist_occ_sites_pre1900 <- left_join(hist_occ_count_pre1900, sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) 

## 1900-1950
hist_occ_1900_1950 <- hist_occ %>%
  filter(year > 1900 & year <= 1950)

### fix the count column
hist_occ_1900_1950 <-  hist_occ_1900_1950 %>%
  mutate(histCount = case_when(individualCount == "X" ~ 1,
                               individualCount == "" ~ 1,
                               TRUE ~ as.numeric(individualCount)))

# calculate total abundance of historical data at each site
hist_occ_count_1900_1950 <- hist_occ_1900_1950 %>%
  group_by(scientific_name, historical_site_code) %>%
  summarise("1900-1950" = sum(histCount)) 

# left join the sites dataframe
hist_occ_sites_1900_1950 <- left_join(hist_occ_count_1900_1950, sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) 

### 1950 - 2000
hist_occ_1950_2000 <- hist_occ %>%
  filter(year > 1950 & year <= 2000)

### fix the count column
hist_occ_1950_2000 <-  hist_occ_1950_2000 %>%
  mutate(histCount = case_when(individualCount == "X" ~ 1,
                               individualCount == "" ~ 1,
                               TRUE ~ as.numeric(individualCount)))

# calculate total abundance of historical data at each site
hist_occ_count_1950_2000 <- hist_occ_1950_2000 %>%
  group_by(scientific_name, historical_site_code) %>%
  summarise("1950-2000" = sum(histCount)) 

# left join the sites dataframe
hist_occ_sites_1950_2000 <- left_join(hist_occ_count_1950_2000, sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) 

## read in modern occurrence data
mod_occ <- read.csv("data/modern-occurrence-data.csv")

## group by scientific name and modern site code
modOcc_spp <- mod_occ %>%
  filter(!is.na(historical_site_code)) %>%
  group_by(scientific_name, modern_site_code) %>%
  summarise("2021" = sum(number))

## left join the sites dataframe
mod_occ_sites <-  left_join(modOcc_spp, sites, by=c("modern_site_code"="modern_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)
```

### Calculate dirichlet distributions of relative abundance

Here, we will functions from Gotelli et al. 2021 to calculate relative species abundances at each site between historic and modern periods
```{r}
## joining the datasets
hist_mod <- full_join(hist_occ_sites_pre1900, hist_occ_sites_1900_1950, by = c("scientific_name","modern_site_code","historical_site_code","elev","site_name","geometry")) %>%
  full_join(., hist_occ_sites_1950_2000, by = c("scientific_name","modern_site_code","historical_site_code","elev","site_name","geometry")) %>%
   full_join(., mod_occ_sites, by = c("scientific_name","modern_site_code","historical_site_code","elev","site_name","geometry")) %>%
  replace(is.na(.), 0) %>%
  arrange(scientific_name)

# note: the level of the analysis is at the 'historic-site' and not the 'modern-site' (which essentially corresponds to three unique sites for every 'historic-site')
# to ensure that the analysis is consistent, we sum abundances across the modern sites corresponding to each historic site

unique_hist_mod <- hist_mod %>%
  group_by(historical_site_code, scientific_name) %>%
  mutate("2021" = sum(`2021`)) 

# remove the modern_site_code, geometry and elev and keep only distinct rows
unique_hist_mod <- distinct(unique_hist_mod[,c(1:3,5,8:10)])

## calculating relative species abundances for the time period 1850-1900
relSpAbund_1850_1900 <- c()

for(i in 1:length(unique(unique_hist_mod$historical_site_code))) {
  
  a <- unique_hist_mod %>%
  filter(historical_site_code == unique(unique_hist_mod$historical_site_code)[i])

  b <- tibble::deframe(a[,c(1,3)]) ## choosing 1850-1900 data
  
  loc <- unique(unique_hist_mod$historical_site_code)[i] # what's the site_code?
  
  relSpAbund_1850_1900[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","1850", sep = ""))

  relSpAbund_1850_1900[[i]] <- relSpAbund_1850_1900[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_1850_1900)[i] <- loc 
}

# save object but full_join across lists
relSpAbund_1850_1900 <- relSpAbund_1850_1900 %>%
    reduce(.f=full_join)

## repeating the above across other time periods
##  time period 1900-1950
relSpAbund_1900_1950 <- c()

for(i in 1:length(unique(unique_hist_mod$historical_site_code))) {
  
  a <- unique_hist_mod %>%
  filter(historical_site_code == unique(unique_hist_mod$historical_site_code)[i])

  b <- tibble::deframe(a[,c(1,5)]) ## choosing 1900-1950 data
  
  loc <- unique(unique_hist_mod$historical_site_code)[i] # what's the site_code?
  
  relSpAbund_1900_1950[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","1900", sep = ""))

  relSpAbund_1900_1950[[i]] <- relSpAbund_1900_1950[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_1900_1950)[i] <- loc 
}

# save object but full_join across lists
relSpAbund_1900_1950 <- relSpAbund_1900_1950 %>%
    reduce(.f=full_join)

##  time period 1950-2000
relSpAbund_1950_2000 <- c()

for(i in 1:length(unique(unique_hist_mod$historical_site_code))) {
  
  a <- unique_hist_mod %>%
  filter(historical_site_code == unique(unique_hist_mod$historical_site_code)[i])

  b <- tibble::deframe(a[,c(1,6)]) ## choosing 1950-2000 data
  
  loc <- unique(unique_hist_mod$historical_site_code)[i] # what's the site_code?
  
  relSpAbund_1950_2000[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","1950", sep = ""))

  relSpAbund_1950_2000[[i]] <- relSpAbund_1950_2000[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_1950_2000)[i] <- loc 
}

# save object but full_join across lists
relSpAbund_1950_2000 <- relSpAbund_1950_2000 %>%
    reduce(.f=full_join)

##  time period 2021
relSpAbund_2021 <- c()

for(i in 1:length(unique(unique_hist_mod$historical_site_code))) {
  
  a <- unique_hist_mod %>%
  filter(historical_site_code == unique(unique_hist_mod$historical_site_code)[i])

  b <- tibble::deframe(a[,c(1,7)]) ## choosing modern data
  
  loc <- unique(unique_hist_mod$historical_site_code)[i] # what's the site_code?
  
  relSpAbund_2021[[i]] <- dirch_stats(b) %>%
  dplyr::select(species, bayes_mean)

  old_varname <-  c('species','bayes_mean')
  new_varname <- c('scientific_name', paste(loc,"_","2021", sep = ""))

  relSpAbund_2021[[i]] <- relSpAbund_2021[[i]] %>%
      data.table::setnames(old = old_varname, new = new_varname)

  names(relSpAbund_2021)[i] <- loc 
}

# save object but full_join across lists
relSpAbund_2021 <- relSpAbund_2021 %>%
    reduce(.f=full_join)

## join all the datasets together
relSpAbund_by_loc <- full_join(relSpAbund_1850_1900, relSpAbund_1900_1950, by = "scientific_name") %>%
  full_join(., relSpAbund_1950_2000, by = "scientific_name") %>%
  full_join(., relSpAbund_2021, by = "scientific_name") %>%
  arrange(scientific_name) %>%
  replace(is.na(.), 0) 
```

## Visualizing differences in relative species abundances over time
```{r}
## pivot longer 
relSpAbund_by_loc <- relSpAbund_by_loc %>%
  pivot_longer(!scientific_name, names_to = "site_year", values_to = "relAbundance") %>%
  separate(site_year, into = c("site", "year"), sep = "_")

## to do:

plot by site - over time
plot by species - over time
(what we need to plot is % change over time?)



```



loc <- unique(unique_hist_mod$historical_site_code)[1]

trial <-  relSpAbund_by_loc %>%
  dplyr::select(scientific_name, contains(loc))



```


