---
editor_options: 
  chunk_output_type: console
---

## Estimating species turnover across time periods

We use temporal diversity metrics from Hallett et al. 2016 based on Collins et al. 2008 and Cleland et al. 2013. We use functions from the package codyn to calculate temporal diversity indices. Please see this link for a detailed vignette: <https://cran.r-project.org/web/packages/codyn/vignettes/Temporal_Diversity_Indices.html>

### Load necessary libraries

```{r}
library(dplyr)
library(stringr)
library(readr)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)
library(codyn)
```

### Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.

```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- extract(alt, sites)
sites <- cbind(sites, elev)
```

For the historical occurrence data, we will choose data from three time periods: 1850-1900, 1900-1950 and 1950-2000.

```{r}
## read in historical occurrence data
hist_occ <- read_csv("data/historical-occurrence-data.csv")

### pre-1900
hist_occ = hist_occ %>%
  mutate(
    year_bin = case_when(
      year <= 1900 ~ "pre1900",
      between(year, 1901, 1950) ~ "1900-1950",
      between(year, 1951, 2000) ~ "1950-2000",
      between(year, 2001, 2020) ~ "2000-2020"
    ),
    histCount = as.numeric(individualCount),
    histCount = if_else(is.na(histCount), 1, histCount)
  ) %>%
  dplyr::select(
    scientific_name, year, year_bin, historical_site_code, histCount
  ) %>%
  group_by(scientific_name, year, year_bin, historical_site_code) %>%
  summarise(histCount = sum(histCount, na.rm = TRUE)) %>%
  ungroup()

## read in modern occurrence data
mod_occ <- read_csv("data/modern-occurrence-data.csv")

## remove sites without historical_site_code
mod_occ <- mod_occ %>%
  filter(!is.na(historical_site_code))

## for the sake of downstream analysis, change the year column
mod_occ$year <- 2021

## subset data
mod_occ <- mod_occ %>%
  group_by(scientific_name, year, historical_site_code) %>%
  summarise(histCount = sum(number)) %>% # rename to histCount
  ungroup() %>%
  mutate(
    year_bin = "2021"
  )
```

## Preparing a dataframe to calculate species turnover

```{r}
# join all data
occ = full_join(hist_occ, mod_occ)

# make a unique identifier for year and site
occ = mutate(
  occ,
  uid = sprintf("%s-%i", historical_site_code, year)
)

# make year bin col, where pre1900 becomes 1850 (for 1850-1900) etc, except 2021
occ = mutate(
  occ,
  year_bin_num = case_when(
    year_bin == "pre1900" ~ 1850,
    year_bin == "1900-1950" ~ 1900,
    year_bin == "1950-2000" ~ 1950,
    year_bin == "2000-2020" ~ 2000,
    year_bin == "2000-2021" ~ 2021
  )
)

# remove locations with only one observation ever
occ = add_count(occ, uid) %>% filter(n > 2)
```

```{r}
## Get unique species abundances 

## Please note: for the sake of downstream analysis, we have renamed 1850-1900 as 1850, 1900-1950 as 1900 and 1950-2000 as 1950. In other words, each year - except for 2021 represents a 50 year time period. 
# 
# ## join all the datasets together
# abund_by_site <- full_join(hist_pre1900, hist_1900_1950, by = c("scientific_name", "historical_site_code", "year")) %>%
#   full_join(., hist_1950_2000, by=c("scientific_name", "historical_site_code", "year")) %>% 
#   full_join(., mod_2021, by=c("scientific_name", "historical_site_code", "year")) %>%
#   arrange(scientific_name) 

# replace NAs to 0
# abund_by_site[,c(4:7)][is.na(abund_by_site)[,c(4:7)]] <- 0
# 
# ## prepare the dataframe
# turnover_dat <-  abund_by_site %>%
#   rowwise() %>%
#   mutate(abundance = sum(`1850`,`1900`,`1950`,`2021`)) %>%
#   ungroup()
# 
# # check structure of dataframes
# turnover_dat <- as.data.frame(turnover_dat)
# turnover_dat$year <- as.integer(turnover_dat$year)
# turnover_dat$scientific_name <- as.factor(turnover_dat$scientific_name)
# turnover_dat$historical_site_code <- as.factor(turnover_dat$historical_site_code)

## We will calculate total turnover as well as the proportion of species that either appear or disappear between time points.

spTurnover <- turnover(df = occ, 
                       time.var = "year", 
                       species.var = "scientific_name", 
                       abundance.var = "histCount",
                       replicate.var = "historical_site_code")


# trial <- turnover(df=knz_001d,
# time.var = "year",
# species.var = "species",
# abundance.var = "abundance",
# replicate.var="subplot")

# to do:
```
