---
editor_options: 
  chunk_output_type: console
---

# Visualizing species persistence probabilities by trait

In this script, we will load species persistence probabilities at the site and landscape level along with species traits to examine how persistence varies over time as a function of species habitat affiliations + site characteristics. 

```{r}
### Load necessary libraries
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)
library(rcompanion)
library(multcomp)
library(report)
library(lme4)
library(sjPlot)
library(ggpubr)
library(glmmTMB)

# source custom functions
source("code/03_persistence-probability-functions.R")

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}
```

## Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

## Load species persistence probabilities at site and landscape level
```{r}
spec_site_persProb <- read.csv( "results/species-persProb-siteLevel.csv")
spec_persProb <- read.csv("results/species-persProb-landscapeLevel.csv")
```

## Load the species persistence probabilities at the site level

We choose data for min and max years to see how historic and modern persistence probabilities vary
```{r}
## select PXt data
siteDat <- spec_site_persProb %>%
  dplyr::select(years,PXt,scientific_name,historical_site_code) %>%
  group_by(scientific_name, historical_site_code) %>%
  filter(years== 2021) %>%
  ungroup()

# visualization of persistence probabilities by site
siteDat$PXt <- as.numeric(siteDat$PXt)

fig_site_persProb <- ggplot(siteDat, aes(x=years, y=PXt, fill = PXt)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico(palette = "roma") +
    theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~historical_site_code)+
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
        legend.position = "none")

ggsave(fig_site_persProb, filename = "figs/fig_persProb_siteLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## while the above figure is useful, we can visualize persistence probability across sites by elevation

siteDat_elev <- left_join(siteDat, sites, by=c("historical_site_code"="historical_site_code"))

fig_persProb_elev <- ggplot(siteDat_elev, aes(x=elev,y=PXt)) +  
  geom_point(shape = 21, colour = "black", fill = "white", size = 2, stroke = 1)+
   geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95,linetype="solid") +  theme_bw() +
  stat_regline_equation(label.y = 0.75, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 0.8, aes(label = ..rr.label..)) +
  labs(x="\nElevation (in meters)", 
       y="Persistence probability\n") +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25))

ggsave(fig_persProb_elev, filename = "figs/fig_persProb_elevation.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Load species persistence probabilities at the landscape level

We choose persistence probabilities for min and max years and repeat analysis we carried out previously. 
```{r}
## select data from the min and max years
specDat <- spec_persProb %>%
  dplyr::select(years,PXt,sciName) %>%
  group_by(sciName) %>%
  filter(years== 2021) %>%
  ungroup()

# visualization of persistence probabilities between min and max year by site
specDat$PXt <- as.numeric(specDat$PXt)

fig_persProb_landscapeLevel <- ggplot(specDat, aes(x=years, y=PXt)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma") +
    theme_bw() +
    labs(x="\nModern persistence probability", 
       y="Persistence probabilty\n") +
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
        legend.position = "none")

ggsave(fig_persProb_landscapeLevel, filename = "figs/fig_persProb_landscapeLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Summarizing persistence probabilities by species trait data

Now that we have persistence probabilities over time, we can examine how they vary for species by site and at the landscape level over time for habitat generalists and specialists. 

#### Let's first curate the species trait data
```{r}
## load state of India's birds data to get habitat affiliation
species_codes <- read.csv("data/species-codes.csv")
habitat_SoIB <- read.csv("data/SoIB-data.csv") # note: we used two files - one file from 2020 and an updated habitat specialization file from 2022 (see below)

# join the two dataframes
habitat_dat <- left_join(species_codes, habitat_SoIB, by=c("scientific_name"="scientific_name"))

# write the above to data folder and recode habitat affiliation
# write.csv(habitat_dat, "data/species-trait-dat.csv", row.names = F)

# note: the classification of habitat type was taken from a working file of trait data for State of India's Birds report v2 (to be released in 2023). We used the following trait classifications:
# Wooded habitat: If it preferred some level of wooded habitat but not particularly specialized on rainforest/forest habitat. An example in this case is the black and orange flycatcher. While it is commonly found in shola forest habitat, this species is also common in degraded wooded areas, including timber plantations. If a species preferred just scrub habitat alone, we lumped it under this category.
# Forest: If a species was a forest specialist. 
# Grassland: While State of India's birds uses a separate classification for grassland & grassland+scrub, here we combine the two and use a single classification for Grassland. Eg. Black drongo - which favors both grassland areas and scrub habitat. There are four species that are classified as using 'open habitat' which had been subsumed under grasslands in our study - this includes - blue-tailed bee-eater, black-winged kite, long-tailed shrike (eBird lists it as preferring open habitats) and western yellow wagtail. 
# Generalist: If a spedcies is truly a generalist species and is not particularly occupying/specialized on any particular wooded/forested/grassland habitat.

## load the species trait data (that includes habitat) and merge with the AVONET dataset
## please note that it is being loaded after curating and adding a column for habitat type based on our understanding of species natural history
hab_dat <- read.csv("data/species-habitat-dat.csv")
avonet <- read.csv("data/AVONET-data.csv")

# join the above two datasets
trait_dat <- left_join(hab_dat, avonet[,c(1,10:20,29:31)], by=c("scientific_name"="Species2"))

## Write the trait data to file
## write.csv(trait_dat,"data/species-trait-dat.csv",row.names = F)
```

## Examining persistence probabilities at the site level in conjunction with species traits
```{r}
## read in trait data (if you have already run the above chunk)
trait_dat <- read.csv("data/species-trait-dat.csv")

## join pers prob and trait datasets
site_dat_trait <- left_join(spec_site_persProb,trait_dat, by=c("scientific_name"="scientific_name"))

## select data from the min and max years
siteDatTrait <- site_dat_trait %>%
  group_by(scientific_name, historical_site_code) %>%
  filter(years== 2021) %>%
  filter(Habitat.type != "Aquatic") %>%
  ungroup()

# visualization of persistence probabilities by site
siteDatTrait$Habitat.type <- factor(siteDatTrait$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))
siteDatTrait$PXt <- as.numeric(siteDatTrait$PXt)

## examining habitat type at site level
fig_site_habitat <- ggplot(siteDatTrait, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +    theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~historical_site_code)+
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      axis.text.x = element_blank(),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14))

ggsave(fig_site_habitat, filename = "figs/fig_persProb_siteLevel_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## examining species persistence prob by trait and elevation

siteDat_elevTrait <- left_join(siteDatTrait, sites, by=c("historical_site_code"="historical_site_code"))

fig_persProb_elevTrait <- ggplot(siteDat_elevTrait, aes(x=elev,y=PXt)) +  
  geom_point(shape = 21, colour = "black", fill = "white", size = 2, stroke = 1)+
   geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95,linetype="solid") +  
  stat_regline_equation(label.y = 1.1, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 1.2, aes(label = ..rr.label..)) +
  theme_bw() +
  facet_wrap(~Habitat.type) +
  labs(x="\nElevation (in meters)", 
       y="Persistence probability\n") +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25))

ggsave(fig_persProb_elevTrait, filename = "figs/fig_persProb_elev_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Examining persistence probabilities at the landscape level in conjunction with species traits
```{r}
## join pers prob and trait datasets
spec_dat_trait <- left_join(spec_persProb,trait_dat, by=c("sciName"="scientific_name"))

## select data from the min and max years
specDatTrait <- spec_dat_trait %>%
  group_by(sciName) %>%
  filter(years== 2021) %>%
  filter(Habitat.type != "Aquatic") %>%
  ungroup()

# visualization of persistence probabilities at the landscape level
specDatTrait$Habitat.type <- factor(specDatTrait$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))
specDatTrait$PXt <- as.numeric(specDatTrait$PXt)

## examining habitat type at site level
fig_spec_habitat <- ggplot(specDatTrait, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +   theme_bw() +
  geom_point(aes(fill = Habitat.type), size = 4, shape = 21, alpha = 0.3,position = position_jitterdodge()) +
    labs(x="", 
       y="Persistence probabilty\n") +
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      axis.text.x = element_blank(),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14))

ggsave(fig_spec_habitat, filename = "figs/fig_persProb_landscapeLevel_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```


## Modeling species persistence probabilities as function of species habitat affiliation and the site habitat type (grouping sites by the modern land cover data)
```{r}
## join and get data ready for plotting
trait_landCover <- left_join(spec_site_persProb, sites, by=c("historical_site_code"="historical_site_code")) %>%
  dplyr::select(years, PXt, scientific_name, modern_landCover_type) %>%
  distinct(.) %>%
  left_join(., trait_dat, by=c("scientific_name"="scientific_name")) %>%
  filter(years==2021) %>%
  filter(Habitat.type != "Aquatic")
  
# visualization of persistence probabilities by site habitat type and species habitat affiliation
trait_landCover$Habitat.type <- factor(trait_landCover$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))

# plot data
## examining habitat type at site level
fig_landCover_trait <- ggplot(trait_landCover, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +  theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~modern_landCover_type)+
  scale_x_discrete(labels = 'Modern') +
    theme(axis.text = element_text(family = "Century Gothic", size = 13),
          axis.text.x = element_blank(),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25)) 

ggsave(fig_landCover_trait, filename = "figs/fig_persProb_habitatType_landCover.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Mixed-effects models (Change in persistence prob ~ species traits)

In this section, we will run mixed-effects models to test the effect of species trait on persistence probability. In other words, has persistence probability for species of a certain habitat type or dispersal ability declined or increased over time?
```{r}
# create a dataframe to run the mixed effects models that join the trait data with the persistence probability datasets

## rerun below with a different persistence probability change dataframe 

changeDat <- left_join(spec_site_persProb,trait_dat[,c(1,5,54,56)], by="scientific_name") %>% mutate(hand_wing_index_z = scale_z(Hand.Wing.Index)) %>%
  mutate(log_weight = log(Mass)) %>%
  mutate(weight_z = scale_z(log_weight)) %>%
  filter(Habitat.type != "Aquatic")

# glmms
changeDat <- changeDat %>%
  mutate(PXt = case_when(PXt == 1 ~ 0.999,
                                  TRUE ~ PXt))

glmm_habType <- glmmTMB(PXt ~ Habitat.type + (1|scientific_name) + (1|historical_site_code), data = changeDat,
                      ziformula = ~1,
                      family = beta_family(link = "logit"))
summary(glmm_habType)

# above model simply suggests that the persistence probability is negatively associated with grassland birds
```
Family: beta  ( logit )
Formula:          
PXt ~ Habitat.type + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

     AIC      BIC   logLik deviance df.resid 
-96814.4 -96740.6  48415.2 -96830.4    75418 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.1099   0.3315  
 historical_site_code (Intercept) 0.1411   0.3756  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 5.34 

Conditional model:
                            Estimate Std. Error z value Pr(>|z|)    
(Intercept)                 1.589825   0.088347  17.995   <2e-16 ***
Habitat.typeGeneralist      0.063096   0.071764   0.879   0.3793    
Habitat.typeGrassland      -0.214470   0.084560  -2.536   0.0112 *  
Habitat.typeWooded Habitat -0.005742   0.064924  -0.088   0.9295    

Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)   -23.34     426.51  -0.055    0.956

```{r}
## Running a model with body mass
glmm_bodyMass <- glmmTMB(PXt ~ weight_z + (1|scientific_name) + (1|historical_site_code), data = changeDat, ziformula = ~1,
                      family = beta_family(link = "logit"))

summary(glmm_bodyMass)

# marginal negative association between body mass and persistence probability
```
Family: beta  ( logit )
Formula:          
PXt ~ weight_z + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

     AIC      BIC   logLik deviance df.resid 
-96812.2 -96756.8  48412.1 -96824.2    75420 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.1127   0.3357  
 historical_site_code (Intercept) 0.1410   0.3755  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 5.34 

Conditional model:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  1.58225    0.07548  20.962   <2e-16 ***
weight_z    -0.04972    0.02186  -2.274    0.023 *  

Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)   -23.39     437.21  -0.054    0.957

```{r}
# dispersal-ability
glmm_hwi <- glmmTMB(PXt ~ hand_wing_index_z + (1|scientific_name) + (1|historical_site_code), data = changeDat, ziformula = ~1,
                      family = beta_family(link = "logit"))
summary(glmm_hwi)

# no significant association
```
 Family: beta  ( logit )
Formula:          
PXt ~ hand_wing_index_z + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

     AIC      BIC   logLik deviance df.resid 
-96810.7 -96755.3  48411.4 -96822.7    75420 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.1137   0.3372  
 historical_site_code (Intercept) 0.1410   0.3755  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 5.34 

Conditional model:
                  Estimate Std. Error z value Pr(>|z|)    
(Intercept)        1.58969    0.07587  20.954   <2e-16 ***
hand_wing_index_z -0.04106    0.02142  -1.917   0.0552 .  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)   -23.22     401.77  -0.058    0.954