---
editor_options: 
  chunk_output_type: console
---

# Visualizing species persistence probabilities by trait

In this script, we will load species persistence probabilities at the site and landscape level along with species traits to examine how persistence varies over time as a function of species habitat affiliations + site characteristics. 

```{r}
### Load necessary libraries
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)
library(rcompanion)
library(multcomp)
library(report)
library(lme4)
library(sjPlot)
library(ggpubr)
library(glmmTMB)

# source custom functions
source("code/03_persistence-probability-functions.R")

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}
```

## Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

## Load species persistence probabilities at site and landscape level
```{r}
spec_site_persProb <- read.csv( "results/species-persProb-siteLevel.csv")
spec_persProb <- read.csv("results/species-persProb-landscapeLevel.csv")
```

## Load the species persistence probabilities at the site level

We choose data for min and max years to see how historic and modern persistence probabilities vary
```{r}
## select PXt data
siteDat <- spec_site_persProb %>%
  dplyr::select(years,PXt,scientific_name,historical_site_code) %>%
  group_by(scientific_name, historical_site_code) %>%
  filter(years== 2021) %>%
  ungroup()

# visualization of persistence probabilities by site
siteDat$PXt <- as.numeric(siteDat$PXt)

fig_site_persProb <- ggplot(siteDat, aes(x=years, y=PXt, fill = PXt)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico(palette = "roma") +
    theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~historical_site_code)+
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
        legend.position = "none")

ggsave(fig_site_persProb, filename = "figs/fig_persProb_siteLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## while the above figure is useful, we can visualize persistence probability across sites by elevation

siteDat_elev <- left_join(siteDat, sites, by=c("historical_site_code"="historical_site_code"))

fig_persProb_elev <- ggplot(siteDat_elev, aes(x=elev,y=PXt)) +  
  geom_point(shape = 21, colour = "black", fill = "white", size = 2, stroke = 1)+
   geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95,linetype="solid") +  theme_bw() +
  stat_regline_equation(label.y = 0.75, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 0.8, aes(label = ..rr.label..)) +
  labs(x="\nElevation (in meters)", 
       y="Persistence probability\n") +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25))

ggsave(fig_persProb_elev, filename = "figs/fig_persProb_elevation.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Load species persistence probabilities at the landscape level

We choose persistence probabilities for min and max years and repeat analysis we carried out previously. 
```{r}
## select data from the min and max years
specDat <- spec_persProb %>%
  dplyr::select(years,PXt,sciName) %>%
  group_by(sciName) %>%
  filter(years== 2021) %>%
  ungroup()

# visualization of persistence probabilities between min and max year by site
specDat$PXt <- as.numeric(specDat$PXt)

fig_persProb_landscapeLevel <- ggplot(specDat, aes(x=years, y=PXt)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma") +
    theme_bw() +
    labs(x="\nModern persistence probability", 
       y="Persistence probabilty\n") +
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
        legend.position = "none")

ggsave(fig_persProb_landscapeLevel, filename = "figs/fig_persProb_landscapeLevel.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Summarizing persistence probabilities by species trait data

Now that we have persistence probabilities over time, we can examine how they vary for species by site and at the landscape level over time for habitat generalists and specialists. 
```{r}
## read in trait data (if you have already run the above chunk)
trait_dat <- read.csv("data/species-trait-dat.csv")

## join pers prob and trait datasets
site_dat_trait <- left_join(spec_site_persProb,trait_dat, by=c("scientific_name"="scientific_name"))

## select data from the min and max years
siteDatTrait <- site_dat_trait %>%
  group_by(scientific_name, historical_site_code) %>%
  filter(years== 2021) %>%
  filter(Habitat.type != "Aquatic") %>%
  ungroup()

# visualization of persistence probabilities by site
siteDatTrait$Habitat.type <- factor(siteDatTrait$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))
siteDatTrait$PXt <- as.numeric(siteDatTrait$PXt)

## examining habitat type at site level
fig_site_habitat <- ggplot(siteDatTrait, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +    theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~historical_site_code)+
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      axis.text.x = element_blank(),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14))

ggsave(fig_site_habitat, filename = "figs/fig_persProb_siteLevel_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

## examining species persistence prob by trait and elevation

siteDat_elevTrait <- left_join(siteDatTrait, sites, by=c("historical_site_code"="historical_site_code"))

fig_persProb_elevTrait <- ggplot(siteDat_elevTrait, aes(x=elev,y=PXt)) +  
  geom_point(shape = 21, colour = "black", fill = "white", size = 2, stroke = 1)+
   geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95,linetype="solid") +  
  stat_regline_equation(label.y = 1.1, aes(label = ..eq.label..)) +
  stat_regline_equation(label.y = 1.2, aes(label = ..rr.label..)) +
  theme_bw() +
  facet_wrap(~Habitat.type) +
  labs(x="\nElevation (in meters)", 
       y="Persistence probability\n") +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25))

ggsave(fig_persProb_elevTrait, filename = "figs/fig_persProb_elev_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Examining persistence probabilities at the landscape level in conjunction with species traits
```{r}
## join pers prob and trait datasets
spec_dat_trait <- left_join(spec_persProb,trait_dat, by=c("sciName"="scientific_name"))

## select data from the min and max years
specDatTrait <- spec_dat_trait %>%
  group_by(sciName) %>%
  filter(years== 2021) %>%
  filter(Habitat.type != "Aquatic") %>%
  ungroup()

# visualization of persistence probabilities at the landscape level
specDatTrait$Habitat.type <- factor(specDatTrait$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))
specDatTrait$PXt <- as.numeric(specDatTrait$PXt)

## examining habitat type at site level
fig_spec_habitat <- ggplot(specDatTrait, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +   theme_bw() +
  geom_point(aes(fill = Habitat.type), size = 4, shape = 21, alpha = 0.3,position = position_jitterdodge()) +
    labs(x="", 
       y="Persistence probabilty\n") +
  scale_x_discrete(labels = 'Modern') +
    theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      axis.text.x = element_blank(),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14))

ggsave(fig_spec_habitat, filename = "figs/fig_persProb_landscapeLevel_habitatType.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```


## Modeling species persistence probabilities as function of species habitat affiliation and the site habitat type (grouping sites by the modern land cover data)
```{r}
## join and get data ready for plotting
trait_landCover <- left_join(spec_site_persProb, sites, by=c("historical_site_code"="historical_site_code")) %>%
  dplyr::select(years, PXt, scientific_name, modern_landCover_type) %>%
  distinct(.) %>%
  left_join(., trait_dat, by=c("scientific_name"="scientific_name")) %>%
  filter(years==2021) %>%
  filter(Habitat.type != "Aquatic")
  
# visualization of persistence probabilities by site habitat type and species habitat affiliation
trait_landCover$Habitat.type <- factor(trait_landCover$Habitat.type, levels=c("Forest", "Grassland", "Generalist", "Wooded Habitat"))

# plot data
## examining habitat type at site level
fig_landCover_trait <- ggplot(trait_landCover, aes(x=years, y=PXt, fill=Habitat.type)) +  geom_boxplot(alpha=0.7) +  
  scale_fill_scico_d(palette = "roma",
                     labels=c("Forest", "Grassland", "Generalist", "Wooded Habitat")) +  theme_bw() +
    labs(x="", 
       y="Persistence probabilty\n") +
  facet_wrap(~modern_landCover_type)+
  scale_x_discrete(labels = 'Modern') +
    theme(axis.text = element_text(family = "Century Gothic", size = 13),
          axis.text.x = element_blank(),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 25)) 

ggsave(fig_landCover_trait, filename = "figs/fig_persProb_habitatType_landCover.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Mixed-effects models (Change in persistence prob ~ species traits)

In this section, we will run mixed-effects models to test the effect of species trait on persistence probability. In other words, has persistence probability for species of a certain habitat type or dispersal ability declined or increased over time?
```{r}
# create a dataframe to run the mixed effects models that join the trait data with the persistence probability datasets

## rerun below with a different persistence probability change dataframe 
changeDat <- left_join(spec_site_persProb,trait_dat[,c(1,5,54,56)], by="scientific_name") %>% mutate(hand_wing_index_z = scale_z(Hand.Wing.Index)) %>%
  mutate(log_weight = log(Mass)) %>%
  mutate(weight_z = scale_z(log_weight)) %>%
  filter(Habitat.type != "Aquatic")

# glmms
changeDat <- changeDat %>%
  mutate(PXt = case_when(PXt == 1 ~ 0.999,
                                  TRUE ~ PXt))

glmm_habType <- glmmTMB(PXt ~ Habitat.type + (1|scientific_name) + (1|historical_site_code), data = changeDat,
                      ziformula = ~1,
                      family = beta_family(link = "logit"))
summary(glmm_habType)

```
Family: beta  ( logit )
Formula:          
PXt ~ Habitat.type + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

      AIC       BIC    logLik  deviance  df.resid 
-728788.8 -728715.0  364402.4 -728804.8     75418 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.1800   0.4243  
 historical_site_code (Intercept) 0.4222   0.6498  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 0.943 

Conditional model:
                           Estimate Std. Error z value Pr(>|z|)    
(Intercept)                -1.25887    0.14019  -8.980   <2e-16 ***
Habitat.typeGeneralist      0.06227    0.09234   0.674   0.5001    
Habitat.typeGrassland      -0.27839    0.10856  -2.564   0.0103 *  
Habitat.typeWooded Habitat -0.03573    0.08342  -0.428   0.6685    

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)   -21.09     138.17  -0.153    0.879

We fitted a zero-inflated logistic mixed model (estimated using ML and
nlminb optimizer) to predict PXt with Habitat.type (formula: PXt ~
Habitat.type). The model included scientific_name as random effects
(formula: list(~1 | scientific_name, ~1 | historical_site_code)). The
model's total explanatory power is substantial (conditional R2 = 0.42) and
the part related to the fixed effects alone (marginal R2) is of 5.87e-03.
The model's intercept, corresponding to Habitat.type = Forest, is at -1.26
(95% CI [-1.53, -0.98], p < .001). Within this model:

  - The effect of Habitat type [Generalist] is statistically non-significant
and positive (beta = 0.06, 95% CI [-0.12, 0.24], p = 0.500; Std. beta =
0.06, 95% CI [-0.12, 0.24])
  - The effect of Habitat type [Grassland] is statistically significant and
negative (beta = -0.28, 95% CI [-0.49, -0.07], p = 0.010; Std. beta =
-0.28, 95% CI [-0.49, -0.07])
  - The effect of Habitat type [Wooded Habitat] is statistically
non-significant and negative (beta = -0.04, 95% CI [-0.20, 0.13], p =
0.668; Std. beta = -0.04, 95% CI [-0.20, 0.13])

Standardized parameters were obtained by fitting the model on a
standardized version of the dataset. 95% Confidence Intervals (CIs) and
p-values were computed using a Wald z-distribution approximation.

```{r}
## Running a model with body mass
glmm_bodyMass <- glmmTMB(PXt ~ weight_z + (1|scientific_name) + (1|historical_site_code), data = changeDat, ziformula = ~1,
                      family = beta_family(link = "logit"))

summary(glmm_bodyMass)
```
Family: beta  ( logit )
Formula:          
PXt ~ weight_z + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

      AIC       BIC    logLik  deviance  df.resid 
-728785.3 -728729.9  364398.7 -728797.3     75420 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.1859   0.4311  
 historical_site_code (Intercept) 0.4219   0.6495  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 0.943 

Conditional model:
            Estimate Std. Error z value Pr(>|z|)    
(Intercept) -1.28735    0.12721 -10.120   <2e-16 ***
weight_z    -0.04938    0.02829  -1.746   0.0809 .  

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)   -21.21     146.65  -0.145    0.885

We fitted a zero-inflated logistic mixed model (estimated using ML and
nlminb optimizer) to predict PXt with weight_z (formula: PXt ~ weight_z).
The model included scientific_name as random effects (formula: list(~1 |
scientific_name, ~1 | historical_site_code)). The model's total explanatory
power is substantial (conditional R2 = 0.42) and the part related to the
fixed effects alone (marginal R2) is of 1.66e-03. The model's intercept,
corresponding to weight_z = 0, is at -1.29 (95% CI [-1.54, -1.04], p <
.001). Within this model:

  - The effect of weight z is statistically non-significant and negative
(beta = -0.05, 95% CI [-0.10, 6.06e-03], p = 0.081; Std. beta = -0.05, 95%
CI [-0.10, 6.07e-03])

Standardized parameters were obtained by fitting the model on a
standardized version of the dataset. 95% Confidence Intervals (CIs) and
p-values were computed using a Wald z-distribution approximation.

```{r}
# dispersal-ability
glmm_hwi <- glmmTMB(PXt ~ hand_wing_index_z + (1|scientific_name) + (1|historical_site_code), data = changeDat, ziformula = ~1,
                      family = beta_family(link = "logit"))
summary(glmm_hwi)
```
 Family: beta  ( logit )
Formula:          
PXt ~ hand_wing_index_z + (1 | scientific_name) + (1 | historical_site_code)
Zero inflation:       ~1
Data: changeDat

      AIC       BIC    logLik  deviance  df.resid 
-728784.0 -728728.6  364398.0 -728796.0     75420 

Random effects:

Conditional model:
 Groups               Name        Variance Std.Dev.
 scientific_name      (Intercept) 0.188    0.4336  
 historical_site_code (Intercept) 0.422    0.6496  
Number of obs: 75426, groups:  
scientific_name, 193; historical_site_code, 28

Dispersion parameter for beta family (): 0.943 

Conditional model:
                  Estimate Std. Error z value Pr(>|z|)    
(Intercept)       -1.28166    0.12758 -10.046   <2e-16 ***
hand_wing_index_z -0.03580    0.02772  -1.291    0.197    

Zero-inflation model:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)    -21.4      161.8  -0.132    0.895

We fitted a zero-inflated logistic mixed model (estimated using ML and
nlminb optimizer) to predict PXt with hand_wing_index_z (formula: PXt ~
hand_wing_index_z). The model included scientific_name as random effects
(formula: list(~1 | scientific_name, ~1 | historical_site_code)). The
model's total explanatory power is substantial (conditional R2 = 0.42) and
the part related to the fixed effects alone (marginal R2) is of 8.74e-04.
The model's intercept, corresponding to hand_wing_index_z = 0, is at -1.28
(95% CI [-1.53, -1.03], p < .001). Within this model:

  - The effect of hand wing index z is statistically non-significant and
negative (beta = -0.04, 95% CI [-0.09, 0.02], p = 0.197; Std. beta = -0.04,
95% CI [-0.09, 0.02])

Standardized parameters were obtained by fitting the model on a
standardized version of the dataset. 95% Confidence Intervals (CIs) and
p-values were computed using a Wald z-distribution approximation.