---
editor_options: 
  chunk_output_type: console
---

## Functional space and hypervolume estimation

In this script, we replicate analyses from Gomez et al. (2021) 
```{r}
### Load necessary libraries
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)
library(pacman)
library(FD)
library(ade4)
library(cowplot)
library(mice)
library(reshape2)
library(hypervolume)
library(alphahull)
library(TTR)
library(plotrix)
library(agricolae)
library(psych)
library(factoextra)
library(ggrepel)
library(tibble)
library(ks)
library(rgl)
library(plotly)

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}

## Function to calculate kernel density probablities
## Date: October 2017
## Outputs: Function named 'cl'
## Args:
## df = dataframe of kernel density data
## prob = Probabilty level e.g. 0.95 (95% confidence level)
## --------------------------------------------------------------

cl <- function(df, prob) {
  dx <- diff(df$x[1:2])
  dy <- diff(df$y[1:2])
  sz <- sort(df$z)
  c1 <- cumsum(sz) * dx * dy
  approx(c1, sz, xout = 1 - prob)$y
}

# function to scale values between 0 and 1
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
```

### Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

## Load species persistence probabilities at site and landscape level
```{r}
spec_site_persProb <- read.csv( "results/species-persProb-siteLevel.csv")
spec_persProb <- read.csv("results/species-persProb-landscapeLevel.csv")
```

### Load species trait data
```{r}
trait_dat <- read.csv("data/species-trait-dat.csv")
```

### Load historical and modern occurrence data

For the historical occurrence data, we will choose data from three time periods: 1850-1900, 1900-1950 and 1950-2000. 
```{r}
## read in historical occurrence data
hist_occ <- read.csv("data/historical-occurrence-data.csv")

### pre-1900
hist_occ_pre1900 <- hist_occ %>%
  filter(year <= 1900)

## group by scientific name and historical site code and join the sites dataframe
hist_occ_pre1900 <- hist_occ_pre1900 %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  rename(., "abundance1850" = n) %>%
  mutate("Y1850" = case_when(abundance1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  left_join(., sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)
  
# repeat above for two more time-periods
## 1900-1950
hist_occ_1900_1950 <- hist_occ %>%
  filter(year > 1900 & year <= 1950)

## group by scientific name and historical site code and join the sites dataframe
hist_occ_1900_1950 <- hist_occ_1900_1950 %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  rename(., "abundance1900" = n) %>%
  mutate("Y1900" = case_when(abundance1900 >= 1 ~ 1,TRUE ~ 0)) %>%
  left_join(., sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)

### 1950 - 2000
hist_occ_1950_2000 <- hist_occ %>%
  filter(year > 1950 & year <= 2000)

## group by scientific name and historical site code and join the sites dataframe
hist_occ_1950_2000 <- hist_occ_1950_2000 %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  rename(., "abundance1950" = n) %>%
  mutate("Y1950" = case_when(abundance1950 >= 1 ~ 1,TRUE ~ 0)) %>%
  left_join(., sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)

## read in modern occurrence data
mod_occ <- read.csv("data/modern-occurrence-data.csv")

## group by scientific name and modern site code
mod_occ <- mod_occ %>%
  filter(!is.na(historical_site_code)) %>%
  group_by(scientific_name, modern_site_code) %>%
  summarise(modCount = sum(number))

mod_occ <-  left_join(mod_occ, sites, by=c("modern_site_code"="modern_site_code")) %>% filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) %>%
  mutate("Y2021" = case_when(modCount >= 1 ~ 1,TRUE ~ 0))
```

## Join the historical and modern count data
```{r}
## joining the historic and modern datasets
## this dataframe can be used to examine land cover data later on
hist_mod <- full_join(hist_occ_pre1900, hist_occ_1900_1950, by = c("scientific_name","modern_site_code","historical_site_code","modern_landCover_type","elev","site_name","geometry")) %>%
  full_join(., hist_occ_1950_2000, by = c("scientific_name","modern_site_code","historical_site_code","modern_landCover_type","elev","site_name","geometry")) %>%
   full_join(., mod_occ, by = c("scientific_name","modern_site_code","historical_site_code","modern_landCover_type","elev","site_name","geometry")) %>%
  replace(is.na(.), 0) %>%
  arrange(scientific_name) %>%
  ungroup()

## Let's use the above dataframe to get species level data for further analysis downstream
## We will ignore site codes

pres_abs <- hist_mod %>%
  dplyr::select(scientific_name, Y1850, Y1900, Y1950, Y2021) %>%
  group_by(scientific_name) %>%
  summarise_all(sum) %>%
  mutate("Y1850" = case_when(Y1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  mutate("Y1900" = case_when(Y1900 >= 1 ~ 1,TRUE ~ 0)) %>%
  mutate("Y1950" = case_when(Y1950 >= 1 ~ 1,TRUE ~ 0)) %>%
  mutate("Y2021" = case_when(Y2021 >= 1 ~ 1,TRUE ~ 0))
```

## Prepare data for a PCA
```{r}
## join morphological trait data with the pres-abs data created
## and subset only columns required for the PCA and next steps
## Here, we will choose scientific_name, habitat type,beak_length_culmen, beak length_nares, beak width, beak depth, tarsus length, wing length, hand wing index, tail length and mass 

trait_pres_abs <-  left_join(pres_abs, trait_dat[,c(1,5,46:51,54:56)], by=c("scientific_name"="scientific_name"))

### add scaled trait measures to traits table
traitScaled <- trait_pres_abs %>% 
  mutate(beak_len_culmen_z = scale_z(Beak.Length_Culmen)) %>%
  mutate(beak_len_nares_z = scale_z(Beak.Length_Nares)) %>%
  mutate(beak_width_z = scale_z(Beak.Width)) %>%
  mutate(beak_depth_z = scale_z(Beak.Depth)) %>%
  mutate(tarsus_len_z = scale_z(Tarsus.Length)) %>%
  mutate(wing_len_z = scale_z(Wing.Length)) %>%
  mutate(hand_wing_index_z = scale_z(Hand.Wing.Index)) %>%
  mutate(tail_len_z = scale_z(Tail.Length)) %>%
  mutate(log_weight = log(Mass)) %>%
  mutate(weight_z = scale_z(log_weight)) %>%
  as.data.frame()

# subset scaled measures
traitScaled <- traitScaled %>% 
  select(scientific_name, Y1850, Y1900, Y1950, Y2021, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z)
 
# table just with columns of morphological traits and species names as rownames
trait_morph_species <- traitScaled %>% 
  select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>%
  column_to_rownames(var = "scientific_name")
```

## Take a PCA of the morphological trait data
```{r}
pcaTotal <- princomp(trait_morph_species, cor = TRUE, scores = TRUE)
summary(pcaTotal)

# scores
scoresPCATotal <- as.data.frame(pcaTotal$scores) %>% 
  tibble::rownames_to_column("scientific_name")

scoresPCATotal <- scoresPCATotal %>% 
  # convert long to wide
  tidyr::gather(key, value, -scientific_name) %>% 
  tidyr::unite(col, key) %>% 
  tidyr::spread(col, value) 

# loadings
loadingsPCATotal <- as.data.frame(unclass(pcaTotal$loadings)) %>% 
  tibble::rownames_to_column("trait") %>% 
  dplyr::bind_rows()

loadingsPCATotal <- loadingsPCATotal %>% 
  # convert long to wide
  tidyr::gather(key, value, -trait) %>% 
  tidyr::unite(col, key) %>% 
  tidyr::spread(col, value) 

# scalar to adjust arrow size
sc_mi <- 7

loadings_sc <- loadingsPCATotal %>% 
  # rescale for arrow sizes
  dplyr::mutate_at(vars(contains("Comp")), funs(.*sc_mi)) %>% 
  # variable names
  mutate(trait = c("Beak Depth","Beak Length (Culmen)","Beak Length (Nares)","Beak Width","Hand-Wing Index", "Tail length", "Tarsus length", "Body mass", "Wing length"))
```

## Join the PCA data with the previously created dataframe of 1s and 0s
```{r}
# subset for species in each period
Y1850 <- traitScaled %>% 
  filter(Y1850 == 1) %>% 
  select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>% 
  left_join(scoresPCATotal, by = "scientific_name") 

Y1900 <- traitScaled %>% 
  filter(Y1900 == 1) %>% 
  select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>% 
  left_join(scoresPCATotal, by = "scientific_name") 

Y1950 <- traitScaled %>% 
  filter(Y1950 == 1) %>% 
  select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>% 
  left_join(scoresPCATotal, by = "scientific_name") 

Y2021 <- traitScaled %>% 
  filter(Y2021 == 1) %>% 
  select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>% 
  left_join(scoresPCATotal, by = "scientific_name")
```

## Multivariate kernel density estimates (following Duong 2019 and Cooke et al. 2019)
```{r}
# kernel density estimation for each period
pc_raw_1850 <- Y1850 %>% 
  # extract first two principal components
  dplyr::select(., scientific_name, Comp.1, Comp.2) %>% 
  tibble::column_to_rownames(var = "scientific_name")

pc_raw_1900 <- Y1900 %>% 
  # extract first two principal components
  dplyr::select(., scientific_name, Comp.1, Comp.2) %>% 
  tibble::column_to_rownames(var = "scientific_name")

pc_raw_1950 <- Y1950 %>% 
  # extract first two principal components
  dplyr::select(., scientific_name, Comp.1, Comp.2) %>% 
  tibble::column_to_rownames(var = "scientific_name")

pc_raw_2021 <- Y2021 %>% 
  # extract first two principal components
  dplyr::select(., scientific_name, Comp.1, Comp.2) %>% 
  tibble::column_to_rownames(var = "scientific_name")

# optimal bandwidth estimation
hpi_1850 <- Hpi(x = pc_raw_1850)
hpi_1900 <- Hpi(x = pc_raw_1900)
hpi_1950 <- Hpi(x = pc_raw_1950)
hpi_2021 <- Hpi(x = pc_raw_2021)

# kernel density estimation 1850  
est_1850 <- kde(x = pc_raw_1850, H = hpi_1850, compute.cont = TRUE)  
den_1850 <- list(est_1850$eval.points[[1]], est_1850$eval.points[[2]], est_1850$estimate)
names(den_1850) <- c("x", "y", "z")
dimnames(den_1850$z) <- list(den_1850$x, den_1850$y)
dcc_1850 <- melt(den_1850$z)

# 0.5 probability kernel
cl_50_1850 <- cl(df = den_1850, prob = 0.50)
# 0.95 probability kernel
cl_95_1850 <- cl(df = den_1850, prob = 0.95)
# 0.99 probability kernel
cl_99_1850 <- cl(df = den_1850, prob = 0.99)

# kernel density estimation 1900   
est_1900 <- kde(x = pc_raw_1900, H = hpi_1900, compute.cont = TRUE)  
den_1900 <- list(est_1900$eval.points[[1]], est_1900$eval.points[[2]], est_1900$estimate)
names(den_1900) <- c("x", "y", "z")
dimnames(den_1900$z) <- list(den_1900$x, den_1900$y)
dcc_1900 <- melt(den_1900$z)

# 0.5 probability kernel
cl_50_1900 <- cl(df = den_1900, prob = 0.50)
# 0.95 probability kernel
cl_95_1900 <- cl(df = den_1900, prob = 0.95)
# 0.99 probability kernel
cl_99_1900 <- cl(df = den_1900, prob = 0.99)

# kernel density estimation 1950   
est_1950 <- kde(x = pc_raw_1950, H = hpi_1950, compute.cont = TRUE)  
den_1950 <- list(est_1950$eval.points[[1]], est_1950$eval.points[[2]], est_1950$estimate)
names(den_1950) <- c("x", "y", "z")
dimnames(den_1950$z) <- list(den_1950$x, den_1950$y)
dcc_1950 <- melt(den_1950$z)

# 0.5 probability kernel
cl_50_1950 <- cl(df = den_1950, prob = 0.50)
# 0.95 probability kernel
cl_95_1950 <- cl(df = den_1950, prob = 0.95)
# 0.99 probability kernel
cl_99_1950 <- cl(df = den_1950, prob = 0.99)

# kernel density estimation 2021  
est_2021 <- kde(x = pc_raw_2021, H = hpi_2021, compute.cont = TRUE)  
den_2021 <- list(est_2021$eval.points[[1]], est_2021$eval.points[[2]], est_2021$estimate)
names(den_2021) <- c("x", "y", "z")
dimnames(den_2021$z) <- list(den_2021$x, den_2021$y)
dcc_2021 <- melt(den_2021$z)

# 0.5 probability kernel
cl_50_2021 <- cl(df = den_2021, prob = 0.50)
# 0.95 probability kernel
cl_95_2021 <- cl(df = den_2021, prob = 0.95)
# 0.99 probability kernel
cl_99_2021 <- cl(df = den_2021, prob = 0.99)

# save principal component data
write.csv(Y1850, file = "data/PCA_1850.csv", row.names = FALSE)
write.csv(Y1900, file = "data/PCA_1900.csv", row.names = FALSE)
write.csv(Y1950, file = "data/PCA_1950.csv", row.names = FALSE)
write.csv(Y2021, file = "data/PCA_2021.csv", row.names = FALSE)
write.csv(scoresPCATotal, file = "data/PCA-scores.csv")
write.csv(loadingsPCATotal, file = "data/PCA-loadings.csv", row.names = FALSE)
```


## Visualize the PCA plots 

By visualizing the PCA of the morphological trait space, we can ask if the functional space has changed over time as a result of changes in community composition
```{r}
# for plots (let's plot each species by it's habitat affiliation on the plot)
wood1850 <- left_join(Y1850, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Wooded Habitat")
for1850 <- left_join(Y1850, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Forest")
grass1850 <- left_join(Y1850, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Grassland")
gen1850 <- left_join(Y1850, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Generalist")

# plot 1850
pca_plot_1850 <- ggplot(dcc_1850, aes(x = Var1, y = Var2)) +
  # coloured probabilty background
  geom_raster(aes(fill = value), show.legend = F) +
  scale_fill_scico(palette = "grayC")+
  # points for species
  geom_point(data = for1850, aes(x = Comp.1, y = Comp.2), size = 2.5, color = "#7E1900", shape = 15) +
  geom_point(data = grass1850, aes(x = Comp.1, y = Comp.2), size = 2.5, colour = "#D9D26A", shape = 16) +
  geom_point(data = gen1850, aes(x = Comp.1, y = Comp.2), size = 2.5, colour = "#60C3D4", shape = 17) +
  geom_point(data = wood1850, aes(x = Comp.1, y = Comp.2), size = 2.5, alpha = 0.5, colour = "#1A3399", shape = 18) +
  # add arrows
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = Comp.1, yend = Comp.2), arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
  # probability kernels
  geom_contour(aes(z = value), breaks = cl_50_1850, colour = "grey30", size = 1) +
  geom_contour(aes(z = value), breaks = cl_95_1850, colour = "grey60", size = 1) +
  geom_contour(aes(z = value), breaks = cl_99_1850, colour = "grey70", size = 1) +
  coord_equal() +
  # add arrows
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = Comp.1, yend = Comp.2), arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
  # add dashed arrows ends
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = -Comp.1, yend = -Comp.2), lty = 5, colour = "darkgrey") +
  # add arrow labels
  geom_text_repel(data = loadings_sc, aes(x = Comp.1, y = Comp.2, label = trait), size = 4, nudge_x = 11, hjust = 0.5, direction = "y", segment.size = 0.5,segment.color = "grey89") +
  scale_x_continuous(limits = c(-5,10), expand = c(0,0)) +
  scale_y_continuous(limits = c(-5,5), expand = c(0,0)) +
  # axis labels - see comp_var
  labs(x = "PC1 - Beak Morphology (67%)", y = "PC2 - Dispersal ability (16%)") +
  theme_classic() +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size =15))

ggsave(pca_plot_1850, filename = "figs/fig_pcaPlot_1850.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)

# for plots (let's plot each species by it's habitat affiliation on the plot)
wood2021 <- left_join(Y2021, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Wooded Habitat")
for2021 <- left_join(Y2021, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Forest")
grass2021 <- left_join(Y2021, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Grassland")
gen2021 <- left_join(Y2021, trait_dat, by="scientific_name") %>%
  filter(Habitat.type == "Generalist")

# plot 2021
pca_plot_2021 <- ggplot(dcc_2021, aes(x = Var1, y = Var2)) +
  # coloured probabilty background
  geom_raster(aes(fill = value), show.legend = F) +
  scale_fill_scico(palette = "grayC")+
  # points for species
  geom_point(data = for2021, aes(x = Comp.1, y = Comp.2), size = 2.5, color = "#7E1900", shape = 15) +
  geom_point(data = grass2021, aes(x = Comp.1, y = Comp.2), size = 2.5, colour = "#D9D26A", shape = 16) +
  geom_point(data = gen2021, aes(x = Comp.1, y = Comp.2), size = 2.5, colour = "#60C3D4", shape = 17) +
  geom_point(data = wood2021, aes(x = Comp.1, y = Comp.2), size = 2.5, alpha = 0.5, colour = "#1A3399", shape = 18) +
  # add arrows
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = Comp.1, yend = Comp.2), arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
  # probability kernels
  geom_contour(aes(z = value), breaks = cl_50_2021, colour = "grey30", size = 1) +
  geom_contour(aes(z = value), breaks = cl_95_2021, colour = "grey60", size = 1) +
  geom_contour(aes(z = value), breaks = cl_99_2021, colour = "grey70", size = 1) +
  coord_equal() +
  # add arrows
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = Comp.1, yend = Comp.2), arrow = arrow(length = unit(0.2, "cm")), colour = "black") +
  # add dashed arrows ends
  geom_segment(data = loadings_sc, aes(x = 0, y = 0, xend = -Comp.1, yend = -Comp.2), lty = 5, colour = "darkgrey") +
  # add arrow labels
  geom_text_repel(data = loadings_sc, aes(x = Comp.1, y = Comp.2, label = trait), size = 4, nudge_x = 11, hjust = 0.5, direction = "y", segment.size = 0.5,segment.color = "grey89") +
  scale_x_continuous(limits = c(-5,10), expand = c(0,0)) +
  scale_y_continuous(limits = c(-5,5), expand = c(0,0)) +
  # axis labels - see comp_var
  labs(x = "PC1 - Beak Morphology (67%)", y = "PC2 - Dispersal ability (16%)") +
  theme_classic() +
  theme(axis.text = element_text(family = "Century Gothic", size = 13),
        legend.title = element_text(family = "Century Gothic"),
    legend.text = element_text(family = "Century Gothic"),
    text = element_text(family = "Century Gothic", size = 15))

ggsave(pca_plot_2021, filename = "figs/fig_pcaPlot_2021.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
```

## Main text Figure 4
```{r}
## patchworking the above two plots for the main text
library(patchwork)

fig_pca_1850_2021 <-
  wrap_plots(pca_plot_1850, pca_plot_2021,
    design = 'AB'
  ) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )

ggsave(fig_pca_1850_2021, filename = "figs/fig04.png", width=15, height=5,device = png(), units="in", dpi = 300); dev.off()
```

## Calculation of hypervolumes
```{r}
# dataframes of trait data (PCA scores)
#Y1850
Y1850A <- Y1850 %>% 
  select(scientific_name, Comp.1, Comp.2, Comp.3, Comp.4, Comp.5, Comp.6, Comp.7, Comp.8, Comp.9) %>% 
  tibble::column_to_rownames(var = "scientific_name")

#Y2021
Y2021A <- Y2021 %>% 
  select(scientific_name, Comp.1, Comp.2, Comp.3, Comp.4, Comp.5, Comp.6, Comp.7, Comp.8, Comp.9) %>% 
  tibble::column_to_rownames(var = "scientific_name")

# hypervolumes: Takes a long time to run.
set.seed(3)
Y1850_hyper_mi <- hypervolume::hypervolume_svm(Y1850A, name = "1850")
saveRDS(Y1850_hyper_mi, "data/1850-hyperVolume-mi9.rds")

set.seed(3)
Y2021_hyper_mi <- hypervolume::hypervolume_svm(Y2021A, name = "2021")
saveRDS(Y2021_hyper_mi, "data/2021-hyper-mi9.rds")

# carry out hypervolume comparisons
# 1850s vs 2021
set.seed(3)
bm_set_mi_1850vs2021 <- hypervolume::hypervolume_set(Y1850_hyper_mi, Y2021_hyper_mi, check.memory = FALSE)

# summarise volumes 1850 vs 2021
sumVolume <- hypervolume::get_volume(bm_set_mi_1850vs2021)

# Plot 3D hypervolumes 
# 1850 vs 2021
bm_set_mi_1850vs2021@HVList$HV1 <- NULL
bm_set_mi_1850vs2021@HVList$HV2 <- NULL
bm_set_mi_1850vs2021@HVList$Union <- NULL

# trait names
names <- c("Beak Morphology", "Dispersal ability","Tarsus length", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9")

colnames(bm_set_mi_1850vs2021@HVList$Intersection@RandomPoints) <- names
colnames(bm_set_mi_1850vs2021@HVList$Unique_1@RandomPoints) <- names
colnames(bm_set_mi_1850vs2021@HVList$Unique_2@RandomPoints) <- names

# plot hypervolumes 
hypervolume::plot.HypervolumeList(bm_set_mi_1850vs2021, show.3d = TRUE, plot.3d.axes.id = c(1,2,3), show.random = FALSE, show.data = TRUE, show.centroid = FALSE, show.density = TRUE, cex.random = 5, colors = c("azure3", "red","cornflowerblue"))
```

## Create fancy 3d plots of hypervolumes
```{r}
# Data frame for Plotly Hyper Volume Plot
inter <- data.frame(bm_set_mi_1850vs2021@HVList$Intersection@RandomPoints)
inter <- inter %>% 
  mutate(section = "intersection")

uni1 <- data.frame(bm_set_mi_1850vs2021@HVList$Unique_1@RandomPoints) 
uni1 <- uni1 %>% 
  mutate(section = "Unique 1850")

uni2 <- data.frame(bm_set_mi_1850vs2021@HVList$Unique_2@RandomPoints)
uni2 <- uni2 %>% 
  mutate(section = "Unique 2021")

HV <- rbind(inter, uni1, uni2)
write.csv(HV,"data/hypervolume-data.csv", row.names=F)

colnames(bm_set_mi_1850vs2021@HVList$Intersection@RandomPoints) <- names
colnames(bm_set_mi_1850vs2021@HVList$Unique_1@RandomPoints) <- names
colnames(bm_set_mi_1850vs2021@HVList$Unique_2@RandomPoints) <- names

pal <- c("green","red", "cornflowerblue" )

fig <- plot_ly(HV, x = ~Beak.Morphology, y = ~(-Dispersal.ability), z = ~(-Tarsus.length), type = 'scatter3d', mode = 'markers', color = ~section, colors = pal, marker = list(size = 3,opacity = 0.6, line = list(color = 'grey66', width = 2)))

fig_hypervolume <- fig %>% layout(scene = list(xaxis = list(title = 'Beak morphology'),yaxis = list(title = 'Dispersal ability'),
             zaxis = list(title = 'Tarsus length')))

fig_hypervolume
```


