---
editor_options: 
  chunk_output_type: console
---

# Functional diversity metrics calculation

This is the second script that replicates analyses from Gomez et al. 2021; here we calculate functional diversity metrics including functional distinctiveness over time.
```{r}
library(factoextra)
library(FD)
library(ks)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyverse)
library(data.table)
library(picante)
library(tibble)
library(iNEXT)
library(funrar)
library(sf)
library(raster)

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}

# function to scale a range of values between 0 and 1
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
```

## Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

### Load species trait data
```{r}
trait_dat <- read.csv("data/species-trait-dat.csv")
```

### Load historical and modern occurrence data

For the historical occurrence data, we will choose data from one time period: 1850-1900
```{r}
## read in historical occurrence data
hist_occ <- read.csv("data/historical-occurrence-data.csv")

### pre-1900
hist_occ <- hist_occ %>%
  filter(year <= 1900)

### fix the count column
hist_occ <-  hist_occ %>%
  mutate(histCount = case_when(individualCount == "X" ~ 1,
                               individualCount == "" ~ 1,
                               TRUE ~ as.numeric(individualCount)))

## remove species with few observations
histCount <- hist_occ %>%
  group_by(scientific_name) %>%
  count() 

## filter species list with a minimum of four records
histCount_minFour <- histCount %>%
  filter(n >= 4 )

## filter species to include only 68 species
hist_occ <- hist_occ %>%
  filter(scientific_name %in% histCount_minFour$scientific_name)

## group by scientific name and historical site code and join the sites dataframe
hist_occ <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  rename(., "abundance1850" = n) %>%
  mutate("Y1850" = case_when(abundance1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  left_join(., sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)
  
## read in modern occurrence data
mod_occ <- read.csv("data/modern-occurrence-data.csv")

## only include species that are present in the historical dataset
mod_occ <- mod_occ %>%
  filter(scientific_name %in% hist_occ$scientific_name)

## group by scientific name and modern site code
mod_occ <- mod_occ %>%
  filter(!is.na(historical_site_code)) %>%
  group_by(scientific_name, modern_site_code) %>%
  summarise(modCount = sum(number))

mod_occ <-  left_join(mod_occ, sites, by=c("modern_site_code"="modern_site_code")) %>% filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) %>%
  mutate("Y2021" = case_when(modCount >= 1 ~ 1,TRUE ~ 0))
```

## Join the historical and modern count data
```{r}
## joining the historic and modern datasets
hist_mod <- full_join(hist_occ, mod_occ, by = c("scientific_name","modern_site_code","historical_site_code","modern_landCover_type","elev","site_name","geometry")) %>%
  replace(is.na(.), 0) %>%
  arrange(scientific_name) %>%
  ungroup()

## Let's use the above dataframe to get species level data for further analysis downstream
## We will ignore site codes
pres_abs <- hist_mod %>%
  dplyr::select(scientific_name, Y1850, Y2021) %>%
  group_by(scientific_name) %>%
  summarise_all(sum) %>%
  mutate("Y1850" = case_when(Y1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  mutate("Y2021" = case_when(Y2021 >= 1 ~ 1,TRUE ~ 0))
```

## Prep dataframes to calculate functional diversity metrics
```{r}
## join trait and presence-absence data
trait_pres_abs <-  left_join(pres_abs, trait_dat[,c(1,5,46:51,54:56)], by=c("scientific_name"="scientific_name"))

### add scaled trait measures to traits table
traitScaled <- trait_pres_abs %>%
  mutate(SD = 1) %>%
  mutate(beak_len_culmen_z = scale_z(Beak.Length_Culmen)) %>%
  mutate(beak_len_nares_z = scale_z(Beak.Length_Nares)) %>%
  mutate(beak_width_z = scale_z(Beak.Width)) %>%
  mutate(beak_depth_z = scale_z(Beak.Depth)) %>%
  mutate(tarsus_len_z = scale_z(Tarsus.Length)) %>%
  mutate(wing_len_z = scale_z(Wing.Length)) %>%
  mutate(hand_wing_index_z = scale_z(Hand.Wing.Index)) %>%
  mutate(tail_len_z = scale_z(Tail.Length)) %>%
  mutate(log_weight = log(Mass)) %>%
  mutate(weight_z = scale_z(log_weight)) %>%
  as.data.frame()

# subset scaled measures
traitScaled <- traitScaled %>% 
  dplyr::select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z)

# order data
traitScaled <- traitScaled[order(traitScaled$scientific_name),]

### prep species year matrices with abundance data

# scaled abundance calculation
species_scaled_abund <- hist_mod %>%
  dplyr::select(scientific_name, abundance1850, modCount) %>%
  group_by(scientific_name) %>%
  summarise_all(sum) %>%
  mutate("1850" = range01(abundance1850)) %>%
  mutate("2021" = range01(modCount)) %>%
  dplyr::select(scientific_name, `1850`, `2021`) %>%
  mutate(sumVar = rowSums(.[2:3])) %>%
  filter(sumVar != 0) %>%
  dplyr::select(scientific_name, `1850`, `2021`) %>%
  pivot_longer(!scientific_name, names_to = "Year", values_to = "Abundance") %>%
  as.data.frame()

# make year column numeric
species_scaled_abund$Year <- as.numeric(species_scaled_abund$Year)

# subset only species with some value of scaled abundance in 1850 or 2021
traitScaledMat <- left_join(species_scaled_abund,traitScaled, by="scientific_name") %>%   dplyr::select(., -Abundance, -Year) %>%
  distinct(.)
traitScaledMat <- traitScaledMat[order(traitScaledMat$scientific_name),]
traitScaledMat <- data.frame(traitScaledMat[,-1], row.names = traitScaledMat[,1])

# create matrix for func. diversity calculations
mat <- species_scaled_abund[order(species_scaled_abund$scientific_name),] %>% 
  pivot_wider(names_from =  scientific_name, values_from = Abundance) %>% 
  column_to_rownames(var = "Year") 

fd_iNDICES <- dbFD(traitScaledMat, mat ,w.abun = TRUE, calc.FRic = TRUE, m = "max", stand.FRic = TRUE , calc.FDiv = TRUE,calc.CWM = FALSE, print.pco = FALSE, messages = TRUE)

write.csv(as.data.frame(fd_iNDICES),"results/observed-func-div-values.csv")

# Estimating Standard Effect Sizes with Null model (takes long time to run)

#__(a)_Unconstrained model_####
#____(i)_calculate FD values for random communities----

mat2 <- species_scaled_abund %>% 
  pivot_wider(names_from =  Year, values_from = Abundance) %>% 
  column_to_rownames(var = "scientific_name")

# mat2 <- as.matrix(mat2)

fdnullall <- NULL
FD0 <- NULL

for(i in 1:999){
  
  randomcom <- randomizeMatrix(t(mat2), null.model='richness',iterations = 100)  
  randomcom <- randomcom[,sort(colnames(randomcom))]
  randomcom <- randomcom[, which(colSums(randomcom) != 0)]

  # extract info from dataframe 
  traitScaledMat <- traitScaled[which((traitScaled$scientific_name %in% colnames(randomcom) ==TRUE)),]
  traitScaledMat <- traitScaledMat[order(traitScaledMat$scientific_name),]
  traitScaledMat <- data.frame(traitScaledMat[,-1], row.names = traitScaledMat[,1])

  fdcom <- data.frame()
  FD0[[i]]   <- dbFD(traitScaledMat, randomcom, w.abun = TRUE, calc.FRic = TRUE, m = "max", stand.FRic = TRUE , calc.FDiv = TRUE,calc.CWM = FALSE, print.pco = FALSE, messages = TRUE) 

  fdcom <- as.data.frame(FD0[[i]])
  fdcom$Year <- rownames(fdcom)
  rownames(fdcom) <- NULL
  fdnullall <- rbind(fdnullall, fdcom)
} 

write.table(fdnullall,'results/FDnullmodel.txt')
```

## Visualizing observed vs. expected values of functional richness and dispersion

```{r}

# Pvalues = (number of Fnull < Fobs + 1)/(999 + 1)
Y1 <- subset(fdnullall, fdnullall$Year == "1850")
Y2 <- subset(fdnullall, fdnullall$Year == "2021")

## FRic Pvalues
ob1 <- nrow(Y1[Y1$FRic< 0.98, ]) + 1
ob2 <- nrow(Y2[Y2$FRic< 0.83, ]) + 1
P1850 <- ob1/(999+1)# P = 0.302
P2021 <- ob2/(999+1)# P = 0.603

## FDis Pvalues
ob1 <- nrow(Y1[Y1$FDis< 2.26, ]) + 1
ob2 <- nrow(Y2[Y2$FDis< 2.93, ]) + 1
P1850 <- ob1/(999+1)# P = 0.087
P2021 <- ob2/(999+1)# P = 0.874

# Boxplots Observed vs Null model
# FRic
fig_obs_v_exp_Fric <- ggplot(fdnullall, aes(x=as.factor(Year), y=FRic)) + 
  geom_boxplot(colour = "gray40", outlier.shape = NA)+ 
  geom_jitter(shape=16, size = 0.1,position=position_jitter(0.1), colour = "gray60") + geom_segment(aes(x = 0.55, y = 0.981477299
, xend = 1.45, yend = 0.981477299
), color = "red", size = 2) +
 geom_segment(aes(x = 1.5, y = 0.839322419
, xend = 2.45, yend = 0.839322419
), color = "red", size = 2)+
  geom_text(x=1, y=1, label="p = 0.302")+
  geom_text(x=2, y=1, label="p = 0.603")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='white', colour = "black"),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20)) +
  labs(x = "", y = "Functional Richness")+ scale_x_discrete(labels = c("1850s", "2021"))

ggsave(fig_obs_v_exp_Fric, filename = "figs/fig_observed_expected_funcRichness.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

# FDis
fig_obs_v_exp_FDis <- ggplot(fdnullall, aes(x=as.factor(Year), y=FDis)) + 
  geom_boxplot(colour = "gray40", outlier.shape = NA)+
  geom_jitter(shape=16, size = 0.1,position=position_jitter(0.1), colour = "gray60") +
  geom_segment(aes(x = 0.55, y = 2.265327387
, xend = 1.45, yend = 2.265327387
), color = "red", size = 2)+
  geom_segment(aes(x = 1.5, y = 2.935667411
, xend = 2.45, yend = 2.935667411
), color = "red", size = 2)+
  geom_text(x=1, y=3.5, label="p = 0.087")+
  geom_text(x=2, y=3.5, label="p = 0.874")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='white', colour = "black"),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20)) +
  labs(x = "", y = "Functional Dispersion")+ scale_x_discrete(labels = c("1850s", "2021"))

ggsave(fig_obs_v_exp_FDis, filename = "figs/fig_observed_expected_funcDispersion.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

## Estimate standard effect sizes

```{r}
# Estimate SES - Standard Effect Size SES = observed mean - mean null / sd null
fd_Mean <- fdnullall %>% 
  group_by(Year) %>%
  summarise_at(vars(FRic, FDis), funs(mean(., na.rm=TRUE))) %>% 
  transmute(Year = Year, FR_M = FRic, FDis_M = FDis)

fd_SES <- fdnullall %>% 
  group_by(Year) %>%
  summarise_at(vars(FRic, FDis), funs(sd(., na.rm=TRUE))) %>% 
  transmute(Year = Year, FR_SD = FRic, FDis_SD = FDis) %>% 
  left_join(fd_Mean, by = "Year") %>% 
  mutate(FRic =c(0.981477299, 0.839322419), FEve = c(0.533492135, 0.393004228), FDiv = c(0.686634956,0.738000497
), FDis = c(2.265327387, 2.935667411
)) %>% 
  mutate(SES_FRIc = ((FRic - FR_M)/FR_SD),SES_FDis = ((FDis - FDis_M)/FDis_SD))

# Plots SES for FRic and FDis
fig_ses_funcRichness <- ggplot(fd_SES, aes(x=as.factor(Year), y=SES_FRIc)) + 
  geom_boxplot(colour = "gray40", outlier.shape = NA) + 
  geom_jitter(shape=16, size = 0.1,position=position_jitter(0.1), colour = "gray60") +  geom_hline(yintercept=0, linetype="dashed", color = "black", size = 1.5) +
  scale_y_continuous(limit = c(-5, 2))+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='white', colour = "black"),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20)) +
  labs(x = "", y = "SES Functional Richness")+ scale_x_discrete(labels = c("1850s", "2021"))

ggsave(fig_ses_funcRichness, filename = "figs/fig_ses_funcRichness.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()

fig_ses_funcDisp <- ggplot(fd_SES, aes(x=as.factor(Year), y=SES_FDis)) + 
  geom_boxplot(colour = "gray40", outlier.shape = NA) + 
  geom_jitter(shape=16, size = 0.1,position=position_jitter(0.1), colour = "gray60") +
  geom_hline(yintercept=0, linetype="dashed", color = "black", size = 1.5)+
  scale_y_continuous(limit = c(-5, 2))+
  
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill='white', colour = "black"),
        axis.text = element_text(colour = "black"),
        legend.position = "none",
        text = element_text(size = 20)) +
  labs(x = "", y = "SES Functional Dispersion")+ scale_x_discrete(labels = c("1850s", "2021"))

ggsave(fig_ses_funcDisp, filename = "figs/fig_ses_funcDispersion.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```

