---
editor_options: 
  chunk_output_type: console
---

# Functional distinctiveness and uniqueness

Here we examine if we have lost functionally distinct species over time

```{r}
## load libraries
library(funrar)
library(dplyr)
library(stringr)
library(tidyverse)
library(scico)
library(RColorBrewer)
library(extrafont)
library(sf)
library(raster)
library(data.table)

# function to z-transform data
scale_z <- function(x){
  (x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
}
```

### Load list of resurvey locations

We will load the list of sites across which a modern resurvey was carried out, along with elevation rasters.
```{r}
# load list of resurvey locations and add elevation as a variable
# remove the modern resurvey locations only (ie. ASFQ)
sites <- read.csv("data/list-of-resurvey-locations.csv")
sites <- sites[-c(1:50),]

# convert to sf object and transform
sites <- st_as_sf(sites, coords = c("longitude", "latitude")) %>%
  `st_crs<-`(4326) %>%
  st_transform(32643)

# add elevation raster
alt <- raster("data/spatial/elevation/alt") # this layer is not added to github as a result of its large size and can be downloaded from SRTM (Farr et al. (2007))

# extract values from that raster (note: transformation of coordinate system)
elev <- raster::extract(alt, sites)
sites <- cbind(sites, elev)
```

### Load species trait data
```{r}
trait_dat <- read.csv("data/species-trait-dat.csv")
```

## Load species relative abundance
```{r}
relAbun <- read.csv("results/species-relative-abundance.csv")
```

### Load historical and modern occurrence data

For the historical occurrence data, we will choose data from a single time periods: 1850-1900
```{r}
## read in historical occurrence data
hist_occ <- read.csv("results/histSubset.csv")

## group by scientific name and historical site code and join the sites dataframe
hist_occ <- hist_occ %>%
  group_by(scientific_name, historical_site_code) %>%
  count() %>%
  rename(., "abundance1850" = n) %>%
  mutate("Y1850" = case_when(abundance1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  left_join(., sites, by=c("historical_site_code"="historical_site_code")) %>%
  filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0)
  
## read in modern occurrence data
mod_occ <- read.csv("results/modSubset.csv")

## group by scientific name and modern site code
mod_occ <- mod_occ %>%
  filter(!is.na(historical_site_code)) %>%
  group_by(scientific_name, modern_site_code) %>%
  summarise(modCount = sum(number))

mod_occ <-  left_join(mod_occ, sites, by=c("modern_site_code"="modern_site_code")) %>% filter(!is.na(scientific_name)) %>%
  replace(is.na(.), 0) %>%
  mutate("Y2021" = case_when(modCount >= 1 ~ 1,TRUE ~ 0))
```

## Join the historical and modern count data
```{r}
## joining the historic and modern datasets
hist_mod <- full_join(hist_occ, mod_occ, by = c("scientific_name","modern_site_code","historical_site_code","modern_landCover_type","elev","site_name","geometry")) %>%
  replace(is.na(.), 0) %>%
  arrange(scientific_name) %>%
  ungroup()

## Let's use the above dataframe to get species level data for further analysis downstream
## We will ignore site codes
pres_abs <- hist_mod %>%
  dplyr::select(scientific_name, Y1850, Y2021) %>%
  group_by(scientific_name) %>%
  summarise_all(sum) %>%
  mutate("Y1850" = case_when(Y1850 >= 1 ~ 1,TRUE ~ 0)) %>%
  mutate("Y2021" = case_when(Y2021 >= 1 ~ 1,TRUE ~ 0))
```

## prep data for functional distinctiveness calculations
```{r}
names(relAbun) <- c("scientific_name","1850-1900","2021")

relAbun <- relAbun %>%
  mutate(relAbunChange = `1850-1900` - `2021`)

## create a matrix
mat <- as.matrix(relAbun)

# subset previous dataframe for list of species that are not detected in 2021 and 1850
# below is being done for the sake of plotting and visualization

notDetected2021 <- pres_abs %>%
  filter(Y2021 == 0, Y1850 == 1) %>%
  mutate("Det/NonDet" = "Naive Extinctions")

notDetected1850 <- pres_abs %>%
  filter(Y1850 == 0, Y2021 == 1) %>%
  mutate("Det/NonDet" = "Novel Colonizations")

shared <- pres_abs %>%
  filter(Y1850 == 1, Y2021 == 1) %>%
  mutate("Det/NonDet" = "Shared")

spDet <- bind_rows(notDetected1850, notDetected2021, shared)

# adding this back to the relAbun dataframe
relAbun <-left_join(relAbun, spDet, by ="scientific_name") 

## moving on to examine the trait data
# scale trait data
traitScaled <- left_join(relAbun, trait_dat[,c(1,5,46:51,54:56)], by ="scientific_name") %>% 
  mutate(beak_len_culmen_z = scale_z(Beak.Length_Culmen)) %>%
  mutate(beak_len_nares_z = scale_z(Beak.Length_Nares)) %>%
  mutate(beak_width_z = scale_z(Beak.Width)) %>%
  mutate(beak_depth_z = scale_z(Beak.Depth)) %>%
  mutate(tarsus_len_z = scale_z(Tarsus.Length)) %>%
  mutate(wing_len_z = scale_z(Wing.Length)) %>%
  mutate(hand_wing_index_z = scale_z(Hand.Wing.Index)) %>%
  mutate(tail_len_z = scale_z(Tail.Length)) %>%
  mutate(log_weight = log(Mass)) %>%
  mutate(weight_z = scale_z(log_weight)) %>%
  as.data.frame() %>%
  filter(Habitat.type != "Aquatic")

# subset scaled measures
traitSubset <- traitScaled %>% 
  dplyr::select(scientific_name, weight_z,
         beak_len_culmen_z, beak_len_nares_z, beak_width_z,
         beak_depth_z, tarsus_len_z, wing_len_z, hand_wing_index_z, tail_len_z) %>%
  column_to_rownames(var = "scientific_name")
```

## Calculate functional distinctiveness
```{r}
#Compute distance matrix
dist_mat <- compute_dist_matrix(traitSubset, metric = "gower")

# Compute distinctiveness for each species
di_df = distinctiveness_stack(com_df = traitScaled,  
                              sp_col = "scientific_name",  # species column
                              com = "Habitat.type",  # habitat.type
                              #abund = "persProbChange",  # persistenceProb column 
                              dist_matrix = dist_mat)  

Di_Sum <- di_df %>% 
  group_by(scientific_name) %>%
  summarise_at(vars(Di), funs(mean(., na.rm=TRUE)))

Di_Sum <- Di_Sum[order(Di_Sum$Di),] %>% 
  mutate(ID = c(1:43)) %>% 
  left_join(relAbun, by = "scientific_name")

# write to file
write.csv(Di_Sum, "results/func-distinctiveness.csv", row.names = F)

## compute functional uniqueness
#Compute Uniqueness for each species
ui_df <- uniqueness_stack(com_df = traitScaled,
                          sp_col = "scientific_name",
                          dist_matrix = dist_mat)

head(ui_df)

Ui_Sum <- ui_df %>% 
  group_by(scientific_name) %>%
  summarise_at(vars(Ui), funs(mean(., na.rm=TRUE)))

Ui_Sum <- Ui_Sum[order(Ui_Sum$Ui),] %>% 
  mutate(ID = c(1:43)) %>% 
  left_join(relAbun, by = "scientific_name")

# visualize/plots
# first we will visualize functional distinctiveness

fig_funcDi_extCol <- ggplot(Di_Sum, aes(x=ID, y=Di, fill = `Det/NonDet`)) + 
  geom_bar(stat = "identity")+ 
  scale_fill_manual(values=c("red", "azure3", "cornflowerblue"), breaks=c("Naive Extinctions","Novel Colonizations","Shared")) +
  theme_bw() +
  geom_text(aes(label = scientific_name, y = 0.3), position = position_dodge(width = 0.9), angle = 90, family = "Century Gothic", fontface = "italic")+
  labs(x = "", y = "Functional distinctiveness - Di") +
  theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14)) +
  guides(fill=guide_legend(title="Legend"))

ggsave(fig_funcDi_extCol, filename = "figs/fig_funcDi_ext_col.png", width = 22, height = 10, device = png(), units = "in", dpi = 300)    
dev.off()

## functional uniqueness
fig_funcUi_extCol <- ggplot(Ui_Sum, aes(x=ID, y=Ui, fill = `Det/NonDet`)) + 
  geom_bar(stat = "identity")+ 
  scale_fill_manual(values=c("red", "azure3", "cornflowerblue"), breaks=c("Naive Extinctions","Novel Colonizations","Shared")) +
  theme_bw() +
  geom_text(aes(label = scientific_name, y = 0.2), position = position_dodge(width = 0.9), angle = 90, family = "Century Gothic", fontface = "italic")+
  labs(x = "", y = "Functional uniqueness - Ui") +
  theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14)) +
  guides(fill=guide_legend(title="Legend"))

ggsave(fig_funcUi_extCol, filename = "figs/fig_funcUi_ext_col.png", width = 22, height = 10, device = png(), units = "in", dpi = 300)    
dev.off()
```

## generalized linear model to test for associations between functional distinctiveness and changes in relative abundance
```{r}
glm_funcDi_relAbun <- glm(relAbunChange ~ Di, data = Di_Sum, family = gaussian(link="identity"))

summary(glm_funcDi_relAbun)
plot_model(glm_funcDi_relAbun, type="pred")
report::report(glm_funcDi_relAbun)
```
Call:
glm(formula = relAbunChange ~ Di, family = gaussian(link = "identity"), 
    data = Di_Sum)

Coefficients:
             Estimate Std. Error t value Pr(>|t|)  
(Intercept)  0.017577   0.007147   2.459   0.0165 *
Di          -0.084283   0.031858  -2.646   0.0102 *

(Dispersion parameter for gaussian family taken to be 0.0004718065)

    Null deviance: 0.034441  on 67  degrees of freedom
Residual deviance: 0.031139  on 66  degrees of freedom
AIC: -323.86

Number of Fisher Scoring iterations: 2

We fitted a linear model (estimated using ML) to predict relAbunChange with
Di (formula: relAbunChange ~ Di). The model's explanatory power is weak (R2
= 0.10). The model's intercept, corresponding to Di = 0, is at 0.02 (95% CI
[3.57e-03, 0.03], t(66) = 2.46, p = 0.014). Within this model:

  - The effect of Di is statistically significant and negative (beta = -0.08,
95% CI [-0.15, -0.02], t(66) = -2.65, p = 0.008; Std. beta = -0.31, 95% CI
[-0.54, -0.08])

Standardized parameters were obtained by fitting the model on a standardized
version of the dataset. 95% Confidence Intervals (CIs) and p-values were
computed using a Wald t-distribution approximation.

## boxplots of functional distinctiveness by det/non-detections
```{r}
fig_funcDi_extCol_boxPlot <- ggplot(Di_Sum, aes(x=`Det/NonDet`, y=Di, fill = `Det/NonDet`)) + 
  geom_boxplot(outlier.shape = NA)+ 
  geom_jitter(shape=16,position=position_jitter(0.2), colour = "black") +
  scale_fill_manual(values=c("red", "cornflowerblue"), breaks=c("Naive Extinctions","Shared")) +
  scale_x_discrete(labels = c("Naive Extinctions","Shared"))+
  theme_bw() +
  labs(x = "", y = "Functional distinctiveness - Di") +
  theme(axis.title = element_text(family = "Century Gothic",
      size = 14, face = "bold"),
        axis.text = element_text(family="Century Gothic",size = 14),
      legend.title = element_text(family="Century Gothic",
                                    size = 14, face = "bold"),
        legend.key.size = unit(1,"cm"),
        legend.text = element_text(family="Century Gothic",size = 14)) +
  guides(fill="none")

ggsave(fig_funcDi_extCol_boxPlot, filename = "figs/fig_funcDi_extCol_boxPlot.png", width = 15, height = 10, device = png(), units = "in", dpi = 300)
dev.off()
```



